---
title: "SIMPlex Breast Cancer — fine-grained cell state analysis and niche characterisation"
subtitle: "Manuscript figures: Fig. 2i-k, Fig. 3, Extended Data Fig. 7-9"
author: "Marcos Tito Machado"
output: html_document
---

```{r config, include=FALSE}
# DATA_ROOT, HDF5 library, palettes — single source of truth for paths.
source(here::here("config.R"))
```

```{r}
# HDF5 library is loaded by config.R
library(semla)
library(tibble)
library(Seurat)
options(Seurat.object.assay.version = "v5")
library(patchwork)
library(plotly)
library(singlet)
library(RcppML)
library(ggplot2)
library(dplyr)
library(tidyr)
library(viridis)
library(pheatmap)
library(cowplot)
library(corrplot)
library(RColorBrewer)
library(heatmap3)
library(harmony)
library(gridExtra)
library(paletteer)
library(Matrix)
```

### Settings 
```{r}
rootObj <- paste0(DATA_ROOT, "/r_objects/breast_cancer/single_nuclei/cross_patient/")
wd <- paste0(FIGS_ROOT, "/breast_cancer/analysis_cellStateLevel/")
if (!dir.exists(rootObj)) {
    dir.create(rootObj, recursive = TRUE)
}
if (!dir.exists(wd)) {
    dir.create(wd, recursive = TRUE)
}

# COLORS
colsCelltype <- rev(paletteer_c("grDevices::Temps", 9))
colsCelltype <- c("#EAE29CFF", "#6CC382FF", "#E99F69FF", "#CF597EFF", "#EAC17AFF", "#29AD8EFF", "#B2D387FF","#E4796DFF", "#089392FF")
colsSample <- paletteer_c("grDevices::Dark 3", 10)
colsSubtype <- c("#6EA9B0", "#B1A6D1", "#E18A96") # rev(paletteer_c("ggthemes::Red-Green-Gold Diverging", 3))
colsMalignancy <- c("#84D7E1","#6ABE8C","#FF95A8")
colsAssays <- c("#6A5ACD", "#FFA500", "#20B2AA")
```

################################################################################# CAF analysis ####################################################################################

#### load progress
```{r}
int.all <- readRDS(paste0(rootObj, "SIMPlex_breast_allSamples.rds"))
samples <- unique(int.all$sample[!grepl("HD", int.all$sample)])
spls <- list()
for (sample in samples) {
    output_path <- paste0(paste0(DATA_ROOT, "/r_objects/breast_cancer/visium/55um/", sample, "/", sample, "_3decon.rds"))
    if (file.exists(output_path)) {
        spls[[sample]] <- readRDS(output_path)
    }
}
```

## Integrate CAFs only 
```{r}
wd <- paste0(FIGS_ROOT, "/breast_cancer/analysis_cellStateLevel/subtype_annotation/")
merged.cafs <- subset(int.all, subset = seuratAnnotation_major == "CAFs")
wd <- paste0(wd, "CAFs/")
if (!dir.exists(wd)) {
    dir.create(wd, recursive = TRUE)
}
```

### remove contaminant cell types and low confidence data
after an initial round of attempted annotation we realize, based on marker genes, that there may be some 
epithelial, immune cell types, or doublets that may skew the results 
```{r}
# Define marker genes and thresholds
positive_markers <- c("COL1A1", "COL1A2")
negative_markers <- c("EPCAM", "KRT8", "CD3D", "CD3E", "MS4A1", "TRAC", "CD163") # immune and epithelial
positive_threshold <- 0.5
negative_threshold <- 0.5

# Fetch expression data for marker genes
expr_pos <- FetchData(merged.cafs, vars = positive_markers)
expr_neg <- FetchData(merged.cafs, vars = negative_markers)

# Identify cells to keep: all positive markers above threshold, all negative markers below threshold
cells_to_keep <- rownames(expr_pos)[
    apply(expr_pos, 1, function(x) all(x > positive_threshold)) &
    apply(expr_neg, 1, function(x) all(x < negative_threshold))
]

cells_before <- ncol(merged.cafs)
merged.cafs <- subset(merged.cafs, cells = cells_to_keep)
cells_after <- ncol(merged.cafs)

cat("Number of cells removed:", cells_before - cells_after, "\n")
```

```{r}
set.seed(1)
merged.cafs <- merged.cafs %>%
  NormalizeData() %>%
  FindVariableFeatures() %>%
  ScaleData() %>%
  RunPCA() %>%
  RunUMAP(dims = 1:30)
# plot pre integration
ggsave(
    filename = paste0(wd, "integration/pca_pre_integration.pdf"),
    plot = DimPlot(merged.cafs, reduction = "pca", group.by = "sample"),
    width = 10,
    height = 10
)
ggsave(
    filename = paste0(wd, "integration/umap_pre_integration.pdf"),
    plot = DimPlot(merged.cafs, reduction = "umap", group.by = c("sample", "patient_ID", "subtype"), ncol = 3),
    width = 30,
    height = 7
)
# integrate
int <- merged.cafs %>% 
    RunHarmony("patient_ID", plot_convergence = TRUE, theta = 0) # min theta since we want to preserve a lot of biological variation
int <- RunUMAP(int, dims = 1:30, reduction = "harmony") 

# plot post integration
ggsave(
    filename = paste0(wd, "integration/pca_post_integration.pdf"),
    plot = DimPlot(int, reduction = "harmony", group.by = "sample"),
    width = 10,
    height = 10
)
ggsave(
    filename = paste0(wd, "integration/umap_post_integration.pdf"),
    plot = DimPlot(int, reduction = "umap", group.by = c("sample", "patient_ID", "subtype"), ncol = 3),
    width = 30,
    height = 7
)
```

### Integrated clustering of CAFS 
```{r}
set.seed(1)
int <- FindNeighbors(int, dims = 1:30, reduction = "harmony")
for (res in c(0.1, 0.25, .3, .4, .5, 0.8, 1, 1.25)){
  int <- FindClusters(int, resolution = res, algorithm = 1, verbose = FALSE)
}

d1 <- DimPlot(int, reduction = "umap", group.by = "RNA_snn_res.0.1", label = TRUE) + 
    DimPlot(int, reduction = "umap", group.by = "RNA_snn_res.0.25", label = TRUE) + 
    DimPlot(int, reduction = "umap", group.by = "RNA_snn_res.0.3", label = TRUE) + 
    DimPlot(int, reduction = "umap", group.by = "RNA_snn_res.0.4", label = TRUE)
d2 <- DimPlot(int, reduction = "umap", group.by = "RNA_snn_res.0.5", label = TRUE) +
    DimPlot(int, reduction = "umap", group.by = "RNA_snn_res.0.8", label = TRUE) + 
    DimPlot(int, reduction = "umap", group.by = "RNA_snn_res.1", label = TRUE) + 
    DimPlot(int, reduction = "umap", group.by = "RNA_snn_res.1.25", label = TRUE)

pdf(paste0(wd, "umap_integratedClusters.pdf"), width = 20, height = 20)
d1 / d2
dev.off()
```

```{r}
renamed <- TRUE
if (renamed == TRUE) {
    int@meta.data$seurat_clusters <- int@meta.data$CAF_major
} else {
    int@meta.data$seurat_clusters <- int@meta.data$RNA_snn_res.0.8
}
Idents(int) <- "seurat_clusters"
ggsave(
    filename = paste0(wd, if (renamed) "renamed_" else "", "umap_post_integration_clusters.pdf"),
    plot = DimPlot(int, reduction = "umap", group.by = c("patient_ID", "subtype", "seurat_clusters"), ncol = 2, label = TRUE, raster = FALSE),
    width = 16,
    height = 12
)
```

## Comparing our CAF clusters to literature

The goal on this section is to compare the subpopulations in https://www.nature.com/articles/s41467-023-39762-1 
and compare it with our CAF cell states.

```{r}

mCAF <- c("FAP","MMP11", "COMP", "POSTN", "COL1A1", "COL1A2", "COL10A1", "COL11A1", "COL8A1", "COL1A2", "COL12A1", "COL3A1", "COL8A1", "COL5A2")
iCAF <- c("CFD", "PLA2G2A", "APOD", "CXCL14", "CD34", "CCLX12", "IL6")
vCAF <- c("NOTCH3", "COL18A1", "MCAM", "MHY11", "RERGL", "ACTA2")
tCAF <- c("PDPN", "MME", "TMEM158", "NDRG1")
ifnCAF <- c("IL32", "CXCL9", "CXCL10", "CXCL11", "IDO1")
apCAF <- c("HLA-DRA", "HLA-DRB1", "CD74")
rCAF <- c("CCL21", "CCL19", "IGFBP5")

caf_celltypes <- list(
  mCAF = mCAF,
  iCAF = iCAF,
  vCAF = vCAF,
  tCAF = tCAF,
  ifnCAF = ifnCAF,
  apCAF = apCAF,
  rCAF = rCAF
)

nat_all_genes <- unique(c(mCAF, iCAF, vCAF, tCAF, ifnCAF, apCAF, rCAF))

Idents(int) <- "seurat_clusters"
p <- DotPlot(int, features = nat_all_genes) + 
        coord_flip() +
        scale_color_gradientn(colors = c("blue", "white", "darkred")) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))
if (renamed) {
    ggsave(paste0(wd, "renamed_nat_markers_dotplot.pdf"), p, width = 10, height = 10)
} else {
    ggsave(paste0(wd, "nat_markers_dotplot.pdf"), p, width = 10, height = 10)
}

ggsave(
    filename = paste0(wd, if (renamed) "renamed_" else "", "nat_markers_umap.pdf"),
    plot = FeaturePlot(int, features = nat_all_genes),
    width = 50,
    height = 70,
    limitsize = FALSE
)
```

#### Gene expression similarity heatmap 

```{r}
nat <- readRDS(paste0(DATA_ROOT, "/external_references/CAF_Cords2023/BREAST_fibro_tumour.rds"))
Idents(int) <- "seurat_clusters"
Idents(nat) <- "CAFtype"
caf_cells <- rownames(int@meta.data)
int <- FindVariableFeatures(int, selection.method = "vst", nfeatures = 3000)
int <- ScaleData(int)

genesNat <- VariableFeatures(nat)
genesCafs <- VariableFeatures(int)

natAvg <- AverageExpression(nat, group.by = "CAFtype", return.seurat = TRUE, assays = "RNA")
natAvg <- ScaleData(natAvg)
natData <- FetchData(object = natAvg, vars = genesNat, layer = "scale.data")
cafsAvg <- AverageExpression(int, group.by = "seurat_clusters", return.seurat = TRUE, assays = "RNA")
cafsAvg <- ScaleData(cafsAvg)
cafsData <- FetchData(object = cafsAvg, vars = genesCafs, layer = "scale.data")

# Extract gene names from each dataset, find common genes, merge datasets
genes_nat <- colnames(natData)
genes_cafs <- colnames(cafsData)
common_genes <- Reduce(intersect, list(genes_nat, genes_cafs))
nat_common <- natData[, common_genes]
cafs_common <- cafsData[, common_genes]

# Calculate correlation between spatial and combined dataset
correlation <- cor(t(nat_common), t(cafs_common))

# Save the plot generated by pheatmap
pdf(paste0(wd, if (renamed) "renamed_" else "", "2023_Cords_natComms_COMPARISON.pdf"), width = 10, height = 10)
heatmap3(correlation, margins = c(13, 10), scale = "none")
dev.off()

print(paste0("Based on ", length(common_genes), " common genes"))
```

## GSEA CAF subpopulations
```{r}
library(singleseqgset)
library(msigdbr)

#### Select Gene Set of interest
# H: hallmark gene sets
# C1: positional gene sets
# C2: curated gene sets

category <- "H" # category from Human MSigDB Collections
h.human <- msigdbr(species="Homo sapiens",category=category)
subcategory <- ""
h.human <- h.human %>% filter(gs_subcat == subcategory)
h.names <- unique(h.human$gs_name)
h.sets <- vector("list",length=length(h.names))
names(h.sets) <- h.names
for (i in names(h.sets)) {
    h.sets[[i]] <- pull(h.human[h.human$gs_name==i,"gene_symbol"])
}
#### Filter out non relevant pathways
# Define the list of pathways to remove
remove_pathways <- c(
  "HALLMARK_BILE_ACID_METABOLISM",         # Liver-specific  
  "HALLMARK_PANCREAS_BETA_CELLS",          # Pancreatic function  
  "HALLMARK_SPERMATOGENESIS",              # Testis-specific  
  "HALLMARK_MYOGENESIS",                   # Muscle development  
  "HALLMARK_ADIPOGENESIS",                 # Fat cell differentiation  
  "HALLMARK_FATTY_ACID_METABOLISM",        # General lipid metabolism  
  "HALLMARK_HEME_METABOLISM",              # Blood-related, not CAF-specific  
  "HALLMARK_CHOLESTEROL_HOMEOSTASIS",      # Unrelated lipid metabolism  
  "HALLMARK_PEROXISOME",                   # General cell detox pathway  
  "HALLMARK_XENOBIOTIC_METABOLISM",        # Detox, not breast cancer-related  
  "HALLMARK_UV_RESPONSE_DN",               # Skin cancer-related  
  "HALLMARK_UV_RESPONSE_UP"                # Skin cancer-related  
)
# Filter the list, keeping only relevant pathways
h.sets <- h.sets[!names(h.sets) %in% remove_pathways]
##### Change names (aesthetic purposes only)
names(h.sets) <- gsub("HALLMARK_", "", names(h.sets))
#### Log Fold change
Idents(int) <- "seurat_clusters"
int$seurat_clusters <- Idents(int)
mtx <- GetAssayData(int, assay = "RNA", layer = "data")
logfc.cafs <- logFC(cluster.ids = int@meta.data$seurat_clusters, expr.mat = mtx)
names(logfc.cafs)
#### Run GSEA
gse.res <- wmw_gsea(expr.mat=mtx, cluster.cells=logfc.cafs[[1]],
                    log.fc.cluster=logfc.cafs[[2]],
                    gene.sets=h.sets)
res.stats <- gse.res[["GSEA_statistics"]]
res.pvals <- gse.res[["GSEA_p_values"]]
res.pvals <- apply(res.pvals,2,p.adjust,method="fdr") #Correct for multiple comparisons
res.stats[order(res.stats[,1],decreasing=TRUE)[1:10],] #Top gene sets enriched by z scores
#Plot the z scores with heatmap3
if (renamed) {
    pdf(paste0(wd, "renamed_GSEA.pdf"), width = 10, height = 12)
} else {
    pdf(paste0(wd, "GSEA.pdf"), width = 10, height = 12)
}
heatmap3(res.stats, margins = c(13, 20))
dev.off()
```

## Rename clusters to putative populations

#### Use Cords 2023 to find rough identities
```{r}
Idents(int) <- "seurat_clusters"
ref <- readRDS(paste0(DATA_ROOT, "/external_references/CAF_Cords2023/BREAST_fibro_tumour.rds"))
ref <- UpdateSeuratObject(ref) # convert v3 to v5
DefaultAssay(ref) <- "RNA"
ref <- ref |>
  NormalizeData() |>
  FindVariableFeatures() |>
  ScaleData()

anchors <- FindTransferAnchors(reference = ref, query = int)
predictions <- TransferData(
  anchorset = anchors,
  refdata = ref$CAFtype,
  verbose = FALSE
)
int <- AddMetaData(object = int, metadata = predictions)
```

#### UMAP and patient distribution bar plot 
```{r}
plot_distribution <- function(obj, metadata_column, fill) {
    # Create a data frame for plotting
    plot_data <- obj@meta.data %>%
        dplyr::count(!!rlang::sym(metadata_column), !!rlang::sym(fill)) %>%
        dplyr::group_by(!!rlang::sym(metadata_column)) %>%
        dplyr::mutate(percentage = n / sum(n) * 100)

    # Calculate entropy for each subpopulation
    entropy <- plot_data %>%
        dplyr::group_by(!!rlang::sym(metadata_column)) %>%
        dplyr::summarize(entropy = -sum(percentage / 100 * log2(percentage / 100)))

    # Merge entropy with plot_data
    plot_data <- plot_data %>%
        dplyr::left_join(entropy, by = metadata_column) %>%
        dplyr::arrange(entropy)

    # Plot the stacked histogram
    p <- ggplot(plot_data, aes(x = reorder(!!rlang::sym(metadata_column), entropy), y = percentage, fill = !!rlang::sym(fill))) +
        geom_bar(stat = "identity") +
        labs(x = "Subpopulations", y = "Percentage", fill = "Metadata") +
        theme_minimal() +
        theme(axis.text.x = element_text(angle = 55, hjust = 1), 
              plot.margin = margin(t = 10,  # Top margin
                                   r = 10,  # Right margin
                                   b = 10,  # Bottom margin
                                   l = 30)) # Left margin
    return(p)
}

p2 <- plot_distribution(int, "seurat_clusters", "patient_ID")
p3 <- plot_distribution(int, "seurat_clusters", "subtype")
p4 <- plot_distribution(int, "seurat_clusters", "predicted.id")
p5 <- p2 + p3 + p4

p1 <- DimPlot(int, group.by = c("seurat_clusters", "predicted.id", "patient_ID", "subtype"), label = TRUE, ncol = 2)
p <- p1 - p5 + plot_layout(ncol = 2, widths = c(5, 5))

ggsave(
    filename = paste0(wd, if (renamed) "renamed_" else "", "summary_plot.pdf"),
    plot = p,
    width = 30,
    height = 10
)
```

### markers
```{r}
Idents(int) <- "seurat_clusters"
markers <- FindAllMarkers(int, assay = "RNA")
top_markers <- markers %>%
  group_by(cluster) %>%
  top_n(n = 5, wt = avg_log2FC)
top100 <- markers %>%
    group_by(cluster) %>%
    top_n(n = 100, wt = avg_log2FC)
write.csv(top100, file = paste0(wd, "caf_markers.csv"), row.names = FALSE)

temp <- subset(int, seurat_clusters %in% na.omit(int$seurat_clusters))
p <- DotPlot(temp, features = unique(top_markers$gene)) + # the subset is to fix a bug related to NAs 
        coord_flip() +
        scale_color_gradientn(colors = c("blue", "white", "darkred")) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))
ggsave(
    filename = paste0(wd, if (renamed) "renamed_" else "", "markers_dotplot.pdf"),
    plot = p,
    width = 10,
    height = 13
) 
```

```{r}
# Violin plot of QC metrics by cluster (no dots)
qc_features <- c("nFeature_RNA", "nCount_RNA", "percent.mt")
if (!"percent.mt" %in% colnames(int@meta.data)) {
    int[["percent.mt"]] <- PercentageFeatureSet(int, pattern = "^MT-")
}
p <- VlnPlot(int, features = qc_features, group.by = "seurat_clusters", pt.size = 0, ncol = 3)
ggsave(
    filename = paste0(wd, "qc_violin_clusters.pdf"),
    plot = p,
    width = 12,
    height = 6
)
```

#### Rename clusters
Renaming clusters based on majority identity of label transfer from Cords et al. 2023 and Patient/Subtype origin

```{r}
int@meta.data$CAF_major <- int@meta.data$seurat_clusters
caf.renames <- c(
    "5" = "apCAF_classical",
    "10" = "apCAF_TcellMimicry",
    "13" = "apCAF_TNBC-enriched",
    "3" = "iCAF_classical",
    "4" = "iCAF_classical",
    "6" = "iCAF_secretory",
    "14" = "iCAF_lipid-associated",
    "11" = "vCAF",
    "1" = "mCAF_plastic",
    "2" = "mCAF_plastic",
    "0" = "mCAF_classical",
    "8" = "mCAF_remodeling",
    "9" = "ifnCAF_interferon",
    "7" = "CAF_unassigned",
    "12" = "mCAF_patient4-specific"
)
int@meta.data$CAF_major <- plyr::mapvalues(int@meta.data$CAF_major, from = names(caf.renames), to = caf.renames)
int@meta.data$seurat_clusters <- int@meta.data$CAF_major
``` 

#### add subclusters back to main object

```{r}
# Create a new metadata column for subpopulation class
int.all$subpopulation <- int.all$seuratAnnotation_major
int.all$subpopulation[names(int$seurat_clusters)] <- as.character(int$seurat_clusters[names(int$seurat_clusters)])
ggsave(
    filename = paste0(wd, "placed_back_CAFclusters.pdf"),
    plot = DimPlot(int.all, group.by = "subpopulation", label = TRUE),
    width = 10,
    height = 8
)
ggsave(
    filename = paste0(wd, "apCAF_sanitycheck.pdf"),
    plot = FeaturePlot(int.all, features = c("CD74", "CIITA", "COL1A1", "COL1A2"), ncol = 2),
    width = 15,
    height = 15
)
```

```{r}
# saveRDS(int, paste0(rootObj, "caf_subpop_analysis.rds"))
# saveRDS(int.all, paste0(rootObj, "integrated_all_subpopulations.rds"))
```

we can go back up and regenerate plots now with the renamed clusters. 
Change renamed to TRUE

### Integrated Spatial mapping 

###### map Vis 55um
```{r}
patient_specific <- FALSE
img_filename <- "patient_combined_subpops_cafs_vis55.jpg"
samples <- unique(int.all$sample[!grepl("HD", int.all$sample)])
spls <- list()
for (sample in samples){
    vis <- readRDS(paste0(paste0(DATA_ROOT, "/r_objects/breast_cancer/visium/55um/", sample, "/", sample, "_spatial.rds"))
    if (patient_specific == TRUE){
        ref <- subset(int.all, subset = sample == !!sample)
    } else {
        ref <- int.all
    }
    

    ## Map w/ semla NNLS
    ident <- "subpopulation"
    celltype <- "subpopulation"
    DefaultAssay(vis) <- "Spatial"
    DefaultAssay(ref) <- "RNA"
    vis <- vis |>
    NormalizeData() |>
    FindVariableFeatures(nfeatures = 5000) |>
    ScaleData()
    ref <- ref |>
    FindVariableFeatures(nfeatures = 2000) # N used here because we want to be specific. The differences between each cell type are very small, so we want to use a smaller number of features
    vis <- RunNNLS(vis, singlecell_object = ref, spatial_assay = DefaultAssay(vis), singlecell_assay = DefaultAssay(ref), groups = ident, slot = "data", assay_name = celltype)
    DefaultAssay(vis) <- celltype

    spls[[sample]] <- vis
}

sample_plots <- list()
for (sample in samples) {
    vis <- spls[[sample]]
    features = unique(gsub("_", "-", int$CAF_major))
    # Ensure all features are present in the vis object
    missing_features <- setdiff(features, rownames(vis))
    vis@assays$subpopulation@data <- rbind(
    vis@assays$subpopulation@data,
    Matrix(0, length(missing_features <- setdiff(missing_features, rownames(vis@assays$subpopulation@data))),
            ncol(vis@assays$subpopulation@data), dimnames = list(missing_features, colnames(vis@assays$subpopulation@data))))

    if (length(vis$sample_id) > 5000) {
        pt_size <- 1
    } else if (length(vis$sample_id) < 5000) {
        pt_size <- 1.6
    }
    p <- MapFeatures(vis, features = features, colors = viridis::turbo(11), label_by = "sample_id", pt_size = pt_size, override_plot_dims = TRUE, scale = "shared", ncol = length(features))
    sample_plots[[sample]] <- p & scale_fill_viridis_c(limits = c(0,1), option = "H")
}
# Merge all sample plots into a final plot
final_plot <- patchwork::wrap_plots(sample_plots, ncol = 1)
ggsave(paste0(wd, "map/", img_filename), final_plot, height = 5*length(samples), width = 3 * length(features), dpi = 300, limitsize = FALSE)
```

################################################################## Epithelial analysis #####################################################################################################

#### load progress
```{r}
int.all <- readRDS(paste0(rootObj, "integrated_all_subpopulations.rds"))
samples <- unique(int.all$sample[!grepl("HD", int.all$sample)])
spls <- list()
for (sample in samples) {
    output_path <- paste0(paste0(DATA_ROOT, "/r_objects/breast_cancer/visium/55um/", sample, "/", sample, "_3decon.rds")
    if (file.exists(output_path)) {
        spls[[sample]] <- readRDS(output_path)
    }
}
```

## Integrate Epithelial only 
```{r}
wd <- paste0(FIGS_ROOT, "/breast_cancer/analysis_cellStateLevel/subtype_annotation/")
wd <- paste0(wd, "Epithelial/")
merged.epi <- subset(int.all, subset = seuratAnnotation_major == "Epithelial")
if (!dir.exists(wd)) {
    dir.create(wd, recursive = TRUE)
}
```

```{r}
set.seed(1)
merged.epi <- merged.epi %>%
  NormalizeData() %>%
  FindVariableFeatures() %>%
  ScaleData() %>%
  RunPCA() %>%
  RunUMAP(dims = 1:30)
# plot pre integration
ggsave(
    filename = paste0(wd, "pca_pre_integration.pdf"),
    plot = DimPlot(merged.epi, reduction = "pca", group.by = "sample"),
    width = 10,
    height = 10
)
ggsave(
    filename = paste0(wd, "umap_pre_integration.pdf"),
    plot = DimPlot(merged.epi, reduction = "umap", group.by = c("sample", "patient_ID", "subtype", "malignancy"), ncol = 2),
    width = 10,
    height = 10
)
# integrate
int <- merged.epi %>% 
    RunHarmony("patient_ID", plot_convergence = TRUE, theta = 0) # min theta since we want to preserve a lot of biological variation
int <- RunUMAP(int, dims = 1:30, reduction = "harmony") 

# plot post integration
ggsave(
    filename = paste0(wd, "pca_post_integration.pdf"),
    plot = DimPlot(int, reduction = "harmony", group.by = "sample"),
    width = 10,
    height = 10
)
ggsave(
    filename = paste0(wd, "umap_post_integration.pdf"),
    plot = DimPlot(int, reduction = "umap", group.by = c("sample", "patient_ID", "subtype", "malignancy"), ncol = 2),
    width = 10,
    height = 10
)
```

### Integrated clustering of Epithelial
```{r}
set.seed(1)
int <- FindNeighbors(int, dims = 1:30)
for (res in c(0.1, 0.25, .3, .4, .5, 0.8, 1, 1.25)){
  int <- FindClusters(int, resolution = res, algorithm = 1, verbose = FALSE)
}

d1 <- DimPlot(int, reduction = "umap", group.by = "RNA_snn_res.0.1", label = TRUE) + 
    DimPlot(int, reduction = "umap", group.by = "RNA_snn_res.0.25", label = TRUE) + 
    DimPlot(int, reduction = "umap", group.by = "RNA_snn_res.0.3", label = TRUE) + 
    DimPlot(int, reduction = "umap", group.by = "RNA_snn_res.0.4", label = TRUE)
d2 <- DimPlot(int, reduction = "umap", group.by = "RNA_snn_res.0.5", label = TRUE) +
    DimPlot(int, reduction = "umap", group.by = "RNA_snn_res.0.8", label = TRUE) + 
    DimPlot(int, reduction = "umap", group.by = "RNA_snn_res.1", label = TRUE) + 
    DimPlot(int, reduction = "umap", group.by = "RNA_snn_res.1.25", label = TRUE)

pdf(paste0(wd, "umap_integratedClusters.pdf"), width = 20, height = 20)
d1 / d2
dev.off()
```

```{r}
renamed <- TRUE
if (renamed == TRUE) {
    int@meta.data$seurat_clusters <- int@meta.data$EPIclusters_renamed
} else {
    int@meta.data$seurat_clusters <- int@meta.data$RNA_snn_res.1
}
Idents(int) <- "seurat_clusters"
ggsave(
    filename = paste0(wd, if (renamed) "renamed_" else "", "umap_post_integration_clusters.pdf"),
    plot = DimPlot(int, reduction = "umap", group.by = c("patient_ID", "subtype", "seurat_clusters"), ncol = 2, label = TRUE, raster = FALSE),
    width = 16,
    height = 12
)
```

## Comparing our EPI clusters to literature

WE compare gene expression to our Reference atlas from BC Atlas

#### Gene expression similarity heatmap 

```{r}
nat <- readRDS(paste0(DATA_ROOT, "/external_references/BC_atlas/miniatlas.rds"))
nat <- subset(nat, subset = celltype_major %in% c("Normal Epithelial", "Cancer Epithelial"))
Idents(int) <- "seurat_clusters"
Idents(nat) <- "celltype_minor"
sub_cells <- rownames(int@meta.data)
int <- FindVariableFeatures(int, selection.method = "vst", nfeatures = 3000)
int <- ScaleData(int)

genesNat <- VariableFeatures(nat)
genesSub <- VariableFeatures(int)

natAvg <- AverageExpression(nat, group.by = "celltype_minor", return.seurat = TRUE, assays = "RNA")
natAvg <- ScaleData(natAvg)
natData <- FetchData(object = natAvg, vars = genesNat, layer = "scale.data")
subAvg <- AverageExpression(int, group.by = "seurat_clusters", return.seurat = TRUE, assays = "RNA")
subAvg <- ScaleData(subAvg)
subData <- FetchData(object = subAvg, vars = genesSub, layer = "scale.data")

# Extract gene names from each dataset, find common genes, merge datasets
genes_nat <- colnames(natData)
genes_sub <- colnames(subData)
common_genes <- Reduce(intersect, list(genes_nat, genes_sub))
nat_common <- natData[, common_genes]
sub_common <- subData[, common_genes]

# Calculate correlation between spatial and combined dataset
correlation <- cor(t(nat_common), t(sub_common))

# Save the plot generated by pheatmap
pdf(paste0(wd, if (renamed) "renamed_" else "", "epithelial_subpop_cluster_comparison_refAtlas.pdf"), width = 10, height = 10)
heatmap3(correlation, margins = c(13, 10), scale = "none")
dev.off()

print(paste0("Based on ", length(common_genes), " common genes"))
```

## GSEA EPI subpopulations
```{r}
library(singleseqgset)
library(msigdbr)

#### Select Gene Set of interest
# H: hallmark gene sets
# C1: positional gene sets
# C2: curated gene sets

category <- "H" # category from Human MSigDB Collections
h.human <- msigdbr(species="Homo sapiens",category=category)
subcategory <- ""
h.human <- h.human %>% filter(gs_subcat == subcategory)
h.names <- unique(h.human$gs_name)
h.sets <- vector("list",length=length(h.names))
names(h.sets) <- h.names
for (i in names(h.sets)) {
    h.sets[[i]] <- pull(h.human[h.human$gs_name==i,"gene_symbol"])
}
#### Filter out non relevant pathways
# Define the list of pathways to remove
remove_pathways <- c(
  "HALLMARK_BILE_ACID_METABOLISM",         # Liver-specific  
  "HALLMARK_PANCREAS_BETA_CELLS",          # Pancreatic function  
  "HALLMARK_SPERMATOGENESIS",              # Testis-specific  
  "HALLMARK_MYOGENESIS",                   # Muscle development  
  "HALLMARK_ADIPOGENESIS",                 # Fat cell differentiation  
  "HALLMARK_FATTY_ACID_METABOLISM",        # General lipid metabolism  
  "HALLMARK_HEME_METABOLISM",              # Blood-related, not CAF-specific  
  "HALLMARK_CHOLESTEROL_HOMEOSTASIS",      # Unrelated lipid metabolism  
  "HALLMARK_PEROXISOME",                   # General cell detox pathway  
  "HALLMARK_XENOBIOTIC_METABOLISM",        # Detox, not breast cancer-related  
  "HALLMARK_UV_RESPONSE_DN",               # Skin cancer-related  
  "HALLMARK_UV_RESPONSE_UP"                # Skin cancer-related  
)
# Filter the list, keeping only relevant pathways
h.sets <- h.sets[!names(h.sets) %in% remove_pathways]
##### Change names (aesthetic purposes only)
names(h.sets) <- gsub("HALLMARK_", "", names(h.sets))
#### Log Fold change
Idents(int) <- "seurat_clusters"
int$seurat_clusters <- Idents(int)
mtx <- GetAssayData(int, assay = "RNA", layer = "data")
logfc.epi <- logFC(cluster.ids = int@meta.data$seurat_clusters, expr.mat = mtx)
names(logfc.epi)
#### Run GSEA
gse.res <- wmw_gsea(expr.mat=mtx, cluster.cells=logfc.epi[[1]],
                    log.fc.cluster=logfc.epi[[2]],
                    gene.sets=h.sets)
res.stats <- gse.res[["GSEA_statistics"]]
res.pvals <- gse.res[["GSEA_p_values"]]
res.pvals <- apply(res.pvals,2,p.adjust,method="fdr") #Correct for multiple comparisons
res.stats[order(res.stats[,1],decreasing=TRUE)[1:10],] #Top gene sets enriched by z scores
#Plot the z scores with heatmap3
if (renamed) {
    pdf(paste0(wd, "renamed_GSEA.pdf"), width = 12, height = 12)
} else {
    pdf(paste0(wd, "GSEA.pdf"), width = 12, height = 12)
}
heatmap3(res.stats, margins = c(13, 20))
dev.off()
```

## Rename clusters to putative populations

#### Use Ref. atlas to compare identities
```{r}
Idents(int) <- "seurat_clusters"
ref <- readRDS(paste0(DATA_ROOT, "/external_references/BC_atlas/miniatlas.rds"))
ref <- subset(ref, subset = celltype_major %in% c("Normal Epithelial", "Cancer Epithelial"))
ref <- UpdateSeuratObject(ref) # convert v3 to v5
DefaultAssay(ref) <- "RNA"
ref <- ref |>
  NormalizeData() |>
  FindVariableFeatures() |>
  ScaleData()

anchors <- FindTransferAnchors(reference = ref, query = int)
predictions <- TransferData(
  anchorset = anchors,
  refdata = ref$celltype_minor,
  verbose = FALSE
)
int <- AddMetaData(object = int, metadata = predictions)
```

#### UMAP and patient distribution bar plot 
```{r}
plot_distribution <- function(obj, metadata_column, fill) {
    # Create a data frame for plotting
    plot_data <- obj@meta.data %>%
        dplyr::count(!!rlang::sym(metadata_column), !!rlang::sym(fill)) %>%
        dplyr::group_by(!!rlang::sym(metadata_column)) %>%
        dplyr::mutate(percentage = n / sum(n) * 100)

    # Calculate entropy for each subpopulation
    entropy <- plot_data %>%
        dplyr::group_by(!!rlang::sym(metadata_column)) %>%
        dplyr::summarize(entropy = -sum(percentage / 100 * log2(percentage / 100)))

    # Merge entropy with plot_data
    plot_data <- plot_data %>%
        dplyr::left_join(entropy, by = metadata_column) %>%
        dplyr::arrange(entropy)

    # Plot the stacked histogram
    p <- ggplot(plot_data, aes(x = reorder(!!rlang::sym(metadata_column), entropy), y = percentage, fill = !!rlang::sym(fill))) +
        geom_bar(stat = "identity") +
        labs(x = "Subpopulations", y = "Percentage", fill = "Metadata") +
        theme_minimal() +
        theme(axis.text.x = element_text(angle = 55, hjust = 1), 
              plot.margin = margin(t = 10,  # Top margin
                                   r = 10,  # Right margin
                                   b = 10,  # Bottom margin
                                   l = 30)) # Left margin
    return(p)
}

p2 <- plot_distribution(int, "seurat_clusters", "patient_ID")
p3 <- plot_distribution(int, "seurat_clusters", "subtype")
p4 <- plot_distribution(int, "seurat_clusters", "predicted.id")
p5 <- p2 + p3 + p4

p1 <- DimPlot(int, group.by = c("seurat_clusters", "predicted.id", "patient_ID", "subtype"), label = TRUE, ncol = 2)
p <- p1 - p5 + plot_layout(ncol = 2, widths = c(5, 7))

ggsave(
    filename = paste0(wd, if (renamed) "renamed_" else "", "summary_plot.pdf"),
    plot = p,
    width = 30,
    height = 10
)
```

### markers
```{r}
Idents(int) <- "seurat_clusters"
markers <- FindAllMarkers(int, assay = "RNA")
top_markers <- markers %>%
  group_by(cluster) %>%
  top_n(n = 5, wt = avg_log2FC)
top100 <- markers %>%
    group_by(cluster) %>%
    top_n(n = 100, wt = avg_log2FC)
write.csv(top100, file = paste0(wd, "epi_markers.csv"), row.names = FALSE)

temp <- subset(int, seurat_clusters %in% na.omit(int$seurat_clusters))
p <- DotPlot(temp, features = unique(top_markers$gene)) + # the subset is to fix a bug related to NAs 
        coord_flip() +
        scale_color_gradientn(colors = c("blue", "white", "darkred")) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))
ggsave(
    filename = paste0(wd, if (renamed) "renamed_" else "", "markers_dotplot.pdf"),
    plot = p,
    width = 10,
    height = 15
) 
```

#### Rename clusters

```{r}
int@meta.data$EPIclusters_renamed <- int@meta.data$seurat_clusters
epi.renames <- c(
    "13" = "LumProgenitors_shared",
    "10" = "Myoepithelial",
    "19" = "LumProgenitors_transitional_pat4",
    "17" = "LumProgenitors_Cycl_pat4", 
    "18" = "Cancer_Basal/Cycl_pat4",
    "2" = "Cancer_Basal/Plastic_pat4", 
    "11" = "Cancer_Transitional_pat4",
    "9" = "Cancer_Cycl_pat2", 
    "24" = "Cancer_Cycl_pat2",
    "5" = "Cancer_LumB_pat2", 
    "16" = "Cancer_LumB_pat2", 
    "21" = "Cancer_Basal/LumB_pat2",
    "22" = "Cancer_Basal_pat9",
    "14" = "Cancer_Basal/LumA_pat1",
    "8" = "Cancer_Lum_pat7",
    "0" = "Cancer_LumA_pat1", 
    "20" = "Cancer_LumA_pat8", 
    "6" = "Cancer_LumA_pat1", 
    "3" = "Cancer_Her2_pat9", 
    "12" = "Cancer_Basal_stem-like_pat6", 
    "4" = "Cancer_Basal_stress_pat6", 
    "23" = "Cancer_Plastic_pat9",
    "7" = "Cancer_Basal_EMT_pat5", 
    "1" = "Cancer_Basal_AntigPres_pat5",
    "15" = "Cancer_Basal_metabolic_pat5"
)
int@meta.data$EPIclusters_renamed <- plyr::mapvalues(int@meta.data$EPIclusters_renamed, from = names(epi.renames), to = epi.renames)
int@meta.data$seurat_clusters <- int@meta.data$EPIclusters_renamed
```


#### add subclusters back to main object

```{r}
# Add to existing subpopulation class metadata
int.all$subpopulation[names(int$seurat_clusters)] <- as.character(int$seurat_clusters[names(int$seurat_clusters)])
```

```{r}
# saveRDS(int, paste0(rootObj, "epithelial_subpop_analysis.rds"))
# saveRDS(int.all, paste0(rootObj, "integrated_all_subpopulations.rds"))
```

we can go back up and regenerate plots now with the renamed clusters. 
Change renamed to TRUE

### Integrated Spatial mapping 

###### map Vis 55um
```{r}
patient_specific <- TRUE
img_filename <- "patient_specific_subpops_epithelial_vis55.jpg"
samples <- unique(int.all$sample[!grepl("HD", int.all$sample)])
spls <- list()
for (sample in samples){
    vis <- readRDS(paste0(paste0(DATA_ROOT, "/r_objects/breast_cancer/visium/55um/", sample, "/", sample, "_spatial.rds"))
    patient_ID <- sub("_.*$", "", sample)
    if (patient_specific == TRUE){
        ref <- subset(int.all, subset = patient_ID == !!patient_ID)
    } else {
        ref <- int.all
    }
    

    ## Map w/ semla NNLS
    ident <- "subpopulation"
    celltype <- "subpopulation"
    DefaultAssay(vis) <- "Spatial"
    DefaultAssay(ref) <- "RNA"
    vis <- vis |>
    NormalizeData() |>
    FindVariableFeatures(nfeatures = 5000) |>
    ScaleData()
    ref <- ref |>
    FindVariableFeatures(nfeatures = 2000) # N used here because we want to be specific. The differences between each cell type are very small, so we want to use a smaller number of features
    vis <- RunNNLS(vis, singlecell_object = ref, spatial_assay = DefaultAssay(vis), singlecell_assay = DefaultAssay(ref), groups = ident, slot = "data", assay_name = celltype)
    DefaultAssay(vis) <- celltype

    spls[[sample]] <- vis
}

sample_plots <- list()
for (sample in samples) {
    vis <- spls[[sample]]
    features = unique(gsub("_", "-", int$EPIclusters_renamed))
    # Ensure all features are present in the vis object
    missing_features <- setdiff(features, rownames(vis))
    vis@assays$subpopulation@data <- rbind(
    vis@assays$subpopulation@data,
    Matrix(0, length(missing_features <- setdiff(missing_features, rownames(vis@assays$subpopulation@data))),
            ncol(vis@assays$subpopulation@data), dimnames = list(missing_features, colnames(vis@assays$subpopulation@data))))

    if (length(vis$sample_id) > 5000) {
        pt_size <- 1
    } else if (length(vis$sample_id) < 5000) {
        pt_size <- 1.6
    }
    p <- MapFeatures(vis, features = features, colors = viridis::turbo(11), label_by = "sample_id", pt_size = pt_size, override_plot_dims = TRUE, scale = "shared", ncol = length(features))
    sample_plots[[sample]] <- p & scale_fill_viridis_c(limits = c(0,1), option = "H")
}
# Merge all sample plots into a final plot
final_plot <- patchwork::wrap_plots(sample_plots, ncol = 1)
ggsave(paste0(wd, "map/", img_filename), final_plot, height = 5*length(samples), width = 3 * length(features), dpi = 300, limitsize = FALSE)
```

################################################################## Immune analysis #####################################################################################################

#### load progress
```{r}
int.all <- readRDS(paste0(rootObj, "integrated_all_subpopulations.rds"))
samples <- unique(int.all$sample[!grepl("HD", int.all$sample)])
spls <- list()
for (sample in samples) {
    output_path <- paste0(paste0(DATA_ROOT, "/r_objects/breast_cancer/visium/55um/", sample, "/", sample, "_3decon.rds")
    if (file.exists(output_path)) {
        spls[[sample]] <- readRDS(output_path)
    }
}
```

## Integrate Immune only 
```{r}
wd <- paste0(FIGS_ROOT, "/breast_cancer/analysis_cellStateLevel/subtype_annotation/")
wd <- paste0(wd, "Immune/")
merged.im <- subset(int.all, subset = seuratAnnotation_major %in% c("T-cells", "B-cells", "Myeloid", "Plasmablasts"))
if (!dir.exists(wd)) {
    dir.create(wd, recursive = TRUE)
}
```

### eliminate potentially mislabelled Epithelial and Fibroblasts
label transfer approach 

```{r}
cells_before <- ncol(merged.im)

ref <- readRDS(paste0(DATA_ROOT, "/external_references/BC_atlas/miniatlas.rds"))
ref <- UpdateSeuratObject(ref) # convert v3 to v5
DefaultAssay(ref) <- "RNA"
ref <- ref |>
  NormalizeData() |>
  FindVariableFeatures() |>
  ScaleData()

anchors <- FindTransferAnchors(reference = ref, query = merged.im)
predictions <- TransferData(
  anchorset = anchors,
  refdata = ref$celltype_major,
  verbose = FALSE
)
merged.im <- AddMetaData(object = merged.im, metadata = predictions)

# Filter out cells where predicted.id is Epithelial or Fibroblasts
merged.im <- subset(merged.im, subset = predicted.id != "CAFs")
merged.im <- subset(merged.im, subset = predicted.id != "Epithelial")

# "final blow" with marker genes
# Remove cells with high expression of EPCAM or COL1A1
merged.im <- subset(merged.im, subset = EPCAM < 1)
merged.im <- subset(merged.im, subset = COL1A1 < 1)

cells_after <- ncol(merged.im)

cat("Number of cells removed:", cells_before - cells_after, "\n")
```

#### integrate
```{r}
set.seed(1)
merged.im <- merged.im %>%
  NormalizeData() %>%
  FindVariableFeatures() %>%
  ScaleData() %>%
  RunPCA() %>%
  RunUMAP(dims = 1:30)
# plot pre integration
ggsave(
    filename = paste0(wd, "pca_pre_integration.pdf"),
    plot = DimPlot(merged.im, reduction = "pca", group.by = "sample"),
    width = 10,
    height = 10
)
ggsave(
    filename = paste0(wd, "umap_pre_integration.pdf"),
    plot = DimPlot(merged.im, reduction = "umap", group.by = c("sample", "patient_ID", "subtype", "seuratAnnotation_major"), ncol = 2),
    width = 30,
    height = 7
)
# integrate
int <- merged.im %>% 
    RunHarmony("patient_ID", plot_convergence = TRUE, theta = 0) # min theta since we want to preserve a lot of biological variation
int <- RunUMAP(int, dims = 1:30, reduction = "harmony") 

# plot post integration
ggsave(
    filename = paste0(wd, "pca_post_integration.pdf"),
    plot = DimPlot(int, reduction = "harmony", group.by = "sample"),
    width = 10,
    height = 10
)
ggsave(
    filename = paste0(wd, "umap_post_integration.pdf"),
    plot = DimPlot(int, reduction = "umap", group.by = c("sample", "patient_ID", "subtype", "seuratAnnotation_major"), ncol = 2),
    width = 15,
    height = 10
)
```

### Integrated clustering of Immune
```{r}
set.seed(1)
int <- FindNeighbors(int, dims = 1:30)
for (res in c(0.1, 0.25, .3, .4, .5, 0.8, 1, 1.25)){
  int <- FindClusters(int, resolution = res, algorithm = 1, verbose = FALSE)
}

d1 <- DimPlot(int, reduction = "umap", group.by = "RNA_snn_res.0.1", label = TRUE) + 
    DimPlot(int, reduction = "umap", group.by = "RNA_snn_res.0.25", label = TRUE) + 
    DimPlot(int, reduction = "umap", group.by = "RNA_snn_res.0.3", label = TRUE) + 
    DimPlot(int, reduction = "umap", group.by = "RNA_snn_res.0.4", label = TRUE)
d2 <- DimPlot(int, reduction = "umap", group.by = "RNA_snn_res.0.5", label = TRUE) +
    DimPlot(int, reduction = "umap", group.by = "RNA_snn_res.0.8", label = TRUE) + 
    DimPlot(int, reduction = "umap", group.by = "RNA_snn_res.1", label = TRUE) + 
    DimPlot(int, reduction = "umap", group.by = "RNA_snn_res.1.25", label = TRUE)

pdf(paste0(wd, "umap_integratedClusters.pdf"), width = 20, height = 20)
d1 / d2
dev.off()
```

```{r}
renamed <- TRUE
if (renamed == TRUE) {
    int@meta.data$seurat_clusters <- int@meta.data$IMclusters_renamed
} else {
    int@meta.data$seurat_clusters <- int@meta.data$RNA_snn_res.0.8
}
Idents(int) <- "seurat_clusters"
ggsave(
    filename = paste0(wd, if (renamed) "renamed_" else "", "umap_post_integration_clusters.pdf"),
    plot = DimPlot(int, reduction = "umap", group.by = c("patient_ID", "subtype", "seurat_clusters"), ncol = 2, label = TRUE, raster = FALSE),
    width = 16,
    height = 12
)
```

## Comparing our Immune clusters to literature

WE compare gene expression to our Reference atlas from BC Atlas (Wu et al. 2021).
The approach used was to select the clustering resolution that allowed the best correlation between clusters and Ref. Atlas "celltype_minor" subpopulations

#### Gene expression similarity heatmap 

```{r}
nat <- readRDS(paste0(DATA_ROOT, "/external_references/BC_atlas/miniatlas.rds"))
nat <- subset(nat, subset = celltype_major %in% c("T-cells", "B-cells", "Myeloid", "Plasmablasts"))
Idents(int) <- "seurat_clusters"
Idents(nat) <- "celltype_minor"
sub_cells <- rownames(int@meta.data)
int <- FindVariableFeatures(int, selection.method = "vst", nfeatures = 3000)
int <- ScaleData(int)

genesNat <- VariableFeatures(nat)
genesSub <- VariableFeatures(int)

natAvg <- AverageExpression(nat, group.by = "celltype_minor", return.seurat = TRUE, assays = "RNA")
natAvg <- ScaleData(natAvg)
natData <- FetchData(object = natAvg, vars = genesNat, layer = "scale.data")
subAvg <- AverageExpression(int, group.by = "seurat_clusters", return.seurat = TRUE, assays = "RNA")
subAvg <- ScaleData(subAvg)
subData <- FetchData(object = subAvg, vars = genesSub, layer = "scale.data")

# Extract gene names from each dataset, find common genes, merge datasets
genes_nat <- colnames(natData)
genes_sub <- colnames(subData)
common_genes <- Reduce(intersect, list(genes_nat, genes_sub))
nat_common <- natData[, common_genes]
sub_common <- subData[, common_genes]

# Calculate correlation between spatial and combined dataset
correlation <- cor(t(nat_common), t(sub_common))

# Save the plot generated by pheatmap
pdf(paste0(wd, if (renamed) "renamed_" else "", "immune_subpop_cluster_comparison_refAtlas.pdf"), width = 10, height = 10)
heatmap3(correlation, margins = c(16, 10), scale = "none")
dev.off()

print(paste0("Based on ", length(common_genes), " common genes"))
```

#### Use Ref. atlas to compare identities
```{r}
Idents(int) <- "seurat_clusters"
ref <- readRDS(paste0(DATA_ROOT, "/external_references/BC_atlas/miniatlas.rds"))
ref <- subset(ref, subset = celltype_major %in% c("T-cells", "B-cells", "Myeloid", "Plasmablasts"))
ref <- UpdateSeuratObject(ref) # convert v3 to v5
DefaultAssay(ref) <- "RNA"
ref <- ref |>
  NormalizeData() |>
  FindVariableFeatures() |>
  ScaleData()

anchors <- FindTransferAnchors(reference = ref, query = int)
predictions <- TransferData(
  anchorset = anchors,
  refdata = ref$celltype_minor,
  verbose = FALSE
)
int <- AddMetaData(object = int, metadata = predictions)
```

#### UMAP and patient distribution bar plot 
```{r}
plot_distribution <- function(obj, metadata_column, fill) {
    # Create a data frame for plotting
    plot_data <- obj@meta.data %>%
        dplyr::count(!!rlang::sym(metadata_column), !!rlang::sym(fill)) %>%
        dplyr::group_by(!!rlang::sym(metadata_column)) %>%
        dplyr::mutate(percentage = n / sum(n) * 100)

    # Calculate entropy for each subpopulation
    entropy <- plot_data %>%
        dplyr::group_by(!!rlang::sym(metadata_column)) %>%
        dplyr::summarize(entropy = -sum(percentage / 100 * log2(percentage / 100)))

    # Merge entropy with plot_data
    plot_data <- plot_data %>%
        dplyr::left_join(entropy, by = metadata_column) %>%
        dplyr::arrange(entropy)

    # Plot the stacked histogram
    p <- ggplot(plot_data, aes(x = reorder(!!rlang::sym(metadata_column), entropy), y = percentage, fill = !!rlang::sym(fill))) +
        geom_bar(stat = "identity") +
        labs(x = "Subpopulations", y = "Percentage", fill = "Metadata") +
        theme_minimal() +
        theme(axis.text.x = element_text(angle = 55, hjust = 1), 
              plot.margin = margin(t = 10,  # Top margin
                                   r = 10,  # Right margin
                                   b = 10,  # Bottom margin
                                   l = 30)) # Left margin
    return(p)
}

p2 <- plot_distribution(int, "seurat_clusters", "patient_ID")
p3 <- plot_distribution(int, "seurat_clusters", "subtype")
p4 <- plot_distribution(int, "seurat_clusters", "predicted.id")
p5 <- p2 + p3 + p4

p1 <- DimPlot(int, group.by = c("seurat_clusters", "predicted.id", "patient_ID", "subtype"), label = TRUE, ncol = 2)
p <- p1 - p5 + plot_layout(ncol = 2, widths = c(5, 5))

ggsave(
    filename = paste0(wd, if (renamed) "renamed_" else "", "summary_plot.pdf"),
    plot = p,
    width = 30,
    height = 10
)
```

### markers
Since the immune clusters are more "delicate" and have very distinctive markers, and the gene correlation plot is ambiguous, lets check the cluster marker genes before renaming
```{r}
Idents(int) <- "seurat_clusters"
im.markers <- FindAllMarkers(int, assay = "RNA")
im.markers %>%
    group_by(cluster) %>%
    top_n(n = 5, wt = avg_log2FC) -> top
im.markers %>%
    group_by(cluster) %>%
    top_n(n = 200, wt = avg_log2FC) -> top100
write.csv(top100, file = paste0(wd, "immune_markers_200.csv"), row.names = FALSE)

# Ensure unique and valid genes
top <- top %>% distinct(gene, .keep_all = TRUE)
valid_genes <- intersect(top$gene, rownames(GetAssayData(int, assay = "RNA")))

p <- DotPlot(int, features = valid_genes) + 
        coord_flip() +
        scale_color_gradientn(colors = c("blue", "white", "darkred")) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))
ggsave(paste0(wd, "cluster_markers_dotplot.pdf"), p, width = 10, height = 14)
```

### plot canonical markers for immune lineages
```{r}
Idents(int) <- "seurat_clusters"

# Pan-immune and lineage-specific immune marker genes
immune_markers <- c(
  # Pan-immune
  "PTPRC", "LYZ", "TYROBP", "CD74", "SPI1",
  
  # Myeloid (macrophages, monocytes, neutrophils)
  "CD14", "CD68", "CSF1R", "ITGAM", "S100A8", "S100A9", "MRC1",
  
  # Dendritic cells
  "CD1C", "CLEC9A", "ITGAX", "BATF3", "IRF8", "LAMP3",
  
  # T cells
  "CD3D", "CD3E", "CD8A", "CD4", "TRAC", "IL7R",
  
  # B cells and plasma cells
  "MS4A1", "CD79A", "CD19", "IGHM", "MZB1", "XBP1",
  
  # NK cells
  "NKG7", "GNLY", "KLRD1", "FCGR3A"
)

# Optional: Add non-immune markers for exclusion
non_immune_markers <- c(
  # Epithelial
  "EPCAM", "KRT8", "KRT18", "KRT5", "CDH1",
  
  # Stromal / fibroblast
  "COL1A1", "COL3A1", "PDGFRA", "DCN", "LUM",
  
  # Endothelial
  "PECAM1", "VWF", "CLDN5", "FLT1"
)

plots <- list()
for (genelist in list(immune_markers, non_immune_markers)) {
    p <- DotPlot(int, features = genelist) + 
        coord_flip() +
        scale_color_gradientn(colors = c("blue", "white", "darkred")) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))
    plots[[paste0("plot_", genelist[1])]] <- p
}
# Combine and save all plots
p3 <- patchwork::wrap_plots(plots, ncol = 1)

p <- p3 + plot_layout(ncol = 3, widths = c(0.6, 0.2, 0.2))
ggsave(paste0(wd, "cluster_markers_canonical.pdf"), p, width = 25, height = 12)

# Also plot UMAPs for the canonical markers, all in one image
umap_plots <- list()
for (genelist in list(immune_markers, non_immune_markers)) {
    for (gene in genelist) {
        if (gene %in% rownames(GetAssayData(int, assay = "RNA"))) {
            p_umap <- FeaturePlot(int, features = gene, reduction = "umap") +
                ggtitle(gene)
            umap_plots[[gene]] <- p_umap
        }
    }
}
# Combine all plots and save as a single PDF
library(patchwork)
all_umaps <- wrap_plots(umap_plots, ncol = 6)
if (renamed) {
    ggsave(
        filename = paste0(wd, "renamed_umap_canonical_markers_all.pdf"),
        plot = all_umaps,
        width = 24,
        height = ceiling(length(umap_plots) / 6) * 4
    )
} else {
    ggsave(
        filename = paste0(wd, "umap_canonical_markers_all.pdf"),
        plot = all_umaps,
        width = 24,
        height = ceiling(length(umap_plots) / 6) * 4
    )
}
```

### Rename clusters

##### CLUSTER 17: Cycling T-cells or Macrophages? 
Plotting genes to get a differential "diagnosis" 
```{r}
# Proliferation markers (shared by cycling cells)
cycling_genes <- c(
  "MKI67", "TOP2A", "PCNA", "CDK1", "CDK2",
  "CDCA3", "CDCA5", "BIRC5", "CENPF", "AURKB",
  "E2F1", "CDT1", "EXO1", "DONSON", "RFC4"
)

# T cell markers
tcell_core <- c("CD3D", "CD3E", "CD3G", "CD2", "CD7", "CD5")
tcell_cd8 <- c("CD8A", "CD8B", "GZMA", "GZMB", "GZMK", "PRF1", "NKG7")
tcell_cd4 <- c("CD4", "IL7R", "LEF1", "CCR7")
tcell_exhaustion <- c("TNFRSF9", "PDCD1", "CTLA4", "LAG3", "HAVCR2", "TIGIT")

tcell_genes <- c(tcell_core, tcell_cd8, tcell_cd4, tcell_exhaustion)

# Macrophage / TAM markers
macrophage_core <- c("CD68", "CD14", "LYZ", "CSF1R")
macrophage_m2 <- c("CD163", "MRC1", "MSR1", "TREM2", "STAB1", "MARCO", "F13A1", "FOLR2")
macrophage_m1 <- c("NOS2", "IL1B", "TNF", "CXCL10")
tam_metabolic <- c("FASN", "EGFR", "SLC2A1", "FABP5", "CD36", "SPP1", "VCAN")

tam_genes <- c(macrophage_core, macrophage_m2, macrophage_m1, tam_metabolic)

plots <- list()
for (genelist in list(cycling_genes, tcell_genes, tam_genes)) {
    p <- DotPlot(int, features = genelist) + 
        coord_flip() +
        scale_color_gradientn(colors = c("blue", "white", "darkred")) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))
    plots[[paste0("plot_", genelist[1])]] <- p
}
# Combine and save all plots
p3 <- patchwork::wrap_plots(plots, ncol = 1)

p <- p3 + plot_layout(ncol = 3, widths = c(1/3, 1/3, 1/3))
ggsave(paste0(wd, "cluster17_cyclingTAM_vs_cyclingTcell.pdf"), p, width = 25, height = 12)
```

CONCLUSION:
    Cluster 17 seems to be clearly cycling T cells given the marker gene expresion.

```{r}
int@meta.data$IMclusters_renamed <- int@meta.data$seurat_clusters
im.renames <- c(
    "0" = "T-cells_atypical", 
    "1" = "T-cells_CD8_effector", 
    "8" = "T-cells_CD4_naive", 
    "4" = "T-cells_CD4_Treg", 
    "5" = "B-cells_activated", 
    "9" = "B/T-mixed_patient4", 
    "10" = "immune_undefined", 
    "12" = "immune_undefined", 
    "19" = "immune_undefined", 
    "17" = "T-cells_cycling", 
    "18" = "Myeloid_atypical_patient6", 
    "21" = "cDC2-like", 
    "20" = "cDC1-like", 
    "22" = "Neutrophils", 
    "3" = "Macrophages_monocyte_derived", 
    "16" = "Macrophages_monocyte_derived",
    "13" = "pDCs",
    "14" = "Mast_cells",
    "6" = "Macrophages_inflammatory",
    "15" = "Macrophages_interferon",
    "2" = "Macrophages_immunoregulatory",
    "11" = "Macrophages_tissue_resident",
    "7" = "Plasma_cells"
)
int@meta.data$IMclusters_renamed <- plyr::mapvalues(int@meta.data$IMclusters_renamed, from = names(im.renames), to = im.renames)
int@meta.data$seurat_clusters <- int@meta.data$IMclusters_renamed
```

### Immune cell state markers
```{r}
Idents(int) <- "seurat_clusters"
im.markers <- FindAllMarkers(int, assay = "RNA")

im.markers %>%
    group_by(cluster) %>%
    top_n(n = 100, wt = avg_log2FC) -> top100
write.csv(top100, file = paste0(wd, "renamed_immune_markers_100.csv"), row.names = FALSE)

top <- top100 %>%
  group_by(cluster) %>%
  slice_head(n = 2) %>%
  ungroup()

p <- DotPlot(int, features = unique(top$gene)) + 
        coord_flip() +
        scale_color_gradientn(colors = c("blue", "white", "darkred")) +
        theme(axis.text.x = element_text(angle = 65, hjust = 1, vjust = 1))
ggsave(paste0(wd, "renamed_markers_dotplot.pdf"), p, width = 8, height = 12)
```

#### add subclusters back to main object

```{r}
# Add to existing subpopulation class metadata
int.all$subpopulation[names(int$seurat_clusters)] <- as.character(int$seurat_clusters[names(int$seurat_clusters)])
```

```{r}
# saveRDS(int, paste0(rootObj, "immune_subpop_analysis.rds"))
# saveRDS(int.all, paste0(rootObj, "integrated_all_subpopulations.rds"))
```

we can go back up and regenerate plots now with the renamed clusters. 
Change renamed to TRUE

## GSEA Immune subpopulations
```{r}
library(singleseqgset)
library(msigdbr)

#### Select Gene Set of interest
# H: hallmark gene sets
# C1: positional gene sets
# C2: curated gene sets

category <- "H" # category from Human MSigDB Collections
h.human <- msigdbr(species="Homo sapiens",category=category)
subcategory <- ""
h.human <- h.human %>% filter(gs_subcat == subcategory)
h.names <- unique(h.human$gs_name)
h.sets <- vector("list",length=length(h.names))
names(h.sets) <- h.names
for (i in names(h.sets)) {
    h.sets[[i]] <- pull(h.human[h.human$gs_name==i,"gene_symbol"])
}
#### Filter out non relevant pathways
# Define the list of pathways to remove
remove_pathways <- c(
  "HALLMARK_BILE_ACID_METABOLISM",         # Liver-specific  
  "HALLMARK_PANCREAS_BETA_CELLS",          # Pancreatic function  
  "HALLMARK_SPERMATOGENESIS",              # Testis-specific  
  "HALLMARK_MYOGENESIS",                   # Muscle development  
  "HALLMARK_ADIPOGENESIS",                 # Fat cell differentiation  
  "HALLMARK_FATTY_ACID_METABOLISM",        # General lipid metabolism  
  "HALLMARK_HEME_METABOLISM",              # Blood-related, not CAF-specific  
  "HALLMARK_CHOLESTEROL_HOMEOSTASIS",      # Unrelated lipid metabolism  
  "HALLMARK_PEROXISOME",                   # General cell detox pathway  
  "HALLMARK_XENOBIOTIC_METABOLISM",        # Detox, not breast cancer-related  
  "HALLMARK_UV_RESPONSE_DN",               # Skin cancer-related  
  "HALLMARK_UV_RESPONSE_UP"                # Skin cancer-related  
)
# Filter the list, keeping only relevant pathways
h.sets <- h.sets[!names(h.sets) %in% remove_pathways]
##### Change names (aesthetic purposes only)
names(h.sets) <- gsub("HALLMARK_", "", names(h.sets))
#### Log Fold change
Idents(int) <- "seurat_clusters"
int$seurat_clusters <- Idents(int)
mtx <- GetAssayData(int, assay = "RNA", layer = "data")
logfc <- logFC(cluster.ids = int@meta.data$seurat_clusters, expr.mat = mtx)
names(logfc)
#### Run GSEA
gse.res <- wmw_gsea(expr.mat=mtx, cluster.cells=logfc[[1]],
                    log.fc.cluster=logfc[[2]],
                    gene.sets=h.sets)
res.stats <- gse.res[["GSEA_statistics"]]
res.pvals <- gse.res[["GSEA_p_values"]]
res.pvals <- apply(res.pvals,2,p.adjust,method="fdr") #Correct for multiple comparisons
res.stats[order(res.stats[,1],decreasing=TRUE)[1:10],] #Top gene sets enriched by z scores
#Plot the z scores with heatmap3
if (renamed) {
    pdf(paste0(wd, "renamed_GSEA.pdf"), width = 12, height = 12)
} else {
    pdf(paste0(wd, "GSEA.pdf"), width = 12, height = 12)
}
heatmap3(res.stats, margins = c(16, 20))
dev.off()
```

### Integrated Spatial mapping 

###### map Vis 55um
```{r}
patient_specific <- TRUE
img_filename <- "patient_specific_subpops_immune_vis55_2000varfeatures.jpg"
samples <- unique(int.all$sample[!grepl("HD", int.all$sample)])
spls <- list()
for (sample in samples){
    vis <- readRDS(paste0(paste0(DATA_ROOT, "/r_objects/breast_cancer/visium/55um/", sample, "/", sample, "_spatial.rds"))
    patient_ID <- sub("_.*$", "", sample)
    if (patient_specific == TRUE){
        ref <- subset(int.all, subset = patient_ID == !!patient_ID)
    } else {
        ref <- int.all
    }
    

    ## Map w/ semla NNLS
    ident <- "subpopulation"
    celltype <- "subpopulation"
    DefaultAssay(vis) <- "Spatial"
    DefaultAssay(ref) <- "RNA"
    vis <- vis |>
    NormalizeData() |>
    FindVariableFeatures(nfeatures = 2000) |>
    ScaleData()
    ref <- ref |>
    FindVariableFeatures(nfeatures = 2000) # N used here because we want to be specific. The differences between each cell type are very small, so we want to use a smaller number of features
    vis <- RunNNLS(vis, singlecell_object = ref, spatial_assay = DefaultAssay(vis), singlecell_assay = DefaultAssay(ref), groups = ident, slot = "data", assay_name = celltype)
    DefaultAssay(vis) <- celltype

    spls[[sample]] <- vis
}

sample_plots <- list()
for (sample in samples) {
    vis <- spls[[sample]]
    features = unique(gsub("_", "-", int$IMclusters_renamed))
    # Ensure all features are present in the vis object
    missing_features <- setdiff(features, rownames(vis))
    vis@assays$subpopulation@data <- rbind(
    vis@assays$subpopulation@data,
    Matrix(0, length(missing_features <- setdiff(missing_features, rownames(vis@assays$subpopulation@data))),
            ncol(vis@assays$subpopulation@data), dimnames = list(missing_features, colnames(vis@assays$subpopulation@data))))

    if (length(vis$sample_id) > 5000) {
        pt_size <- 1
    } else if (length(vis$sample_id) < 5000) {
        pt_size <- 1.6
    }
    p <- MapFeatures(vis, features = features, colors = viridis::turbo(11), label_by = "sample_id", pt_size = pt_size, override_plot_dims = TRUE, scale = "shared", ncol = length(features))
    sample_plots[[sample]] <- p & scale_fill_viridis_c(limits = c(0,1), option = "H")
}
# Merge all sample plots into a final plot
final_plot <- patchwork::wrap_plots(sample_plots, ncol = 1)
ggsave(paste0(wd, "map/", img_filename), final_plot, height = 5*length(samples), width = 3 * length(features), dpi = 300, limitsize = FALSE)
```

################################################################## Epithelial, Immune and CAF co-analysis ##############################################################################

# Settings
```{r}
wd <- paste0(FIGS_ROOT, "/breast_cancer/analysis_cellStateLevel/final_subpopulations/")
```

## load 
```{r}
int.all <- readRDS(paste0(rootObj, "integrated_all_subpopulations.rds"))
mergedVis <- readRDS(paste0(rootObj, "merged_visium_deconSubpopulations_simplex.rds"))
```

### final cleanup 
remove the celltypes before that have been deemed low confidence in the initial annotation 

```{r}
int.all <- subset(int.all, subpopulation != "B-cells")
int.all <- subset(int.all, subpopulation != "CAFs")
int.all <- subset(int.all, subpopulation != "T-cells")
int.all <- subset(int.all, subpopulation != "Myeloid")
int.all <- subset(int.all, subpopulation != "Plasmablasts")
int.all <- subset(int.all, subpopulation != "immune_undefined")
int.all <- subset(int.all, subpopulation != "CAF_unassigned")
int.all <- subset(int.all, subpopulation != "apCAF_TNBC-enriched") # downstream analysis showed these were low confidence: possible plasma cell contamination
```

```{r}
ggsave(
    filename = paste0(wd, "umaps/all_subpopulations_UMAP.pdf"),
    plot = DimPlot(int.all, reduction = "umap", group.by = c("subpopulation", "sample", "subtype"), ncol = 1, label = TRUE, raster = FALSE, repel = TRUE),
    width = 15,
    height = 20
)
```

## Spatial correlation subpopulations

###### map Vis 55um
here we map with patient_specific to be more biologically accurate and with patient_specific = FALSE if we want to see 
which populations map more patient specifically 

```{r}
patient_specific <- TRUE 
img_filename <- "patSpec_allSubpops_vis55_2000varfeatures.jpg"
samples <- unique(int.all$sample[!grepl("HD", int.all$sample)])
spls <- list()
for (sample in samples){
    vis <- readRDS(paste0(paste0(DATA_ROOT, "/r_objects/breast_cancer/visium/55um/", sample, "/", sample, "_spatial.rds"))
    patient_ID <- sub("_.*$", "", sample)
    if (patient_specific == TRUE){
        ref <- subset(int.all, subset = patient_ID == !!patient_ID)
    } else {
        ref <- int.all
    }
    
    ## Map w/ semla NNLS
    ident <- "subpopulation"
    celltype <- "subpopulation"
    DefaultAssay(vis) <- "Spatial"
    DefaultAssay(ref) <- "RNA"
    vis <- vis |>
    NormalizeData() |>
    FindVariableFeatures(nfeatures = 2000) |>
    ScaleData()
    ref <- ref |>
    FindVariableFeatures(nfeatures = 2000) # N used here because we want to be specific. The differences between each cell type are very small, so we want to use a smaller number of features
    vis <- RunNNLS(vis, singlecell_object = ref, spatial_assay = DefaultAssay(vis), singlecell_assay = DefaultAssay(ref), groups = ident, slot = "data", assay_name = celltype)
    DefaultAssay(vis) <- celltype

    spls[[sample]] <- vis
}

features = unique(gsub("_", "-", int.all$subpopulation))
for (sample in samples) {
    vis <- spls[[sample]]
    # Ensure all features are present in the vis object
    missing_features <- setdiff(features, rownames(vis))
    vis@assays$subpopulation@data <- rbind(
    vis@assays$subpopulation@data,
    Matrix(0.0, length(missing_features <- setdiff(missing_features, rownames(vis@assays$subpopulation@data))),
            ncol(vis@assays$subpopulation@data), dimnames = list(missing_features, colnames(vis@assays$subpopulation@data))))

    # save
    spls[[sample]] <- vis

    # if (length(vis$sample_id) > 5000) {
    #     pt_size <- 0.6
    # } else if (length(vis$sample_id) < 5000) {
    #     pt_size <- 1
    # }
    # p <- MapFeatures(vis, features = features, colors = viridis::turbo(11), label_by = "sample_id", pt_size = pt_size, override_plot_dims = TRUE, scale = "shared", ncol = 10)
    # p <- p & scale_fill_viridis_c(limits = c(0,1), option = "H") # make scale 0 to 1
    # ggsave(
    #     filename = paste0(wd, "map_decon/", sample, "_", img_filename),
    #     plot = p,
    #     height = 3*6,
    #     width = 3*8,
    #     dpi = 300,
    #     limitsize = FALSE
    # )
}
```

###### plot each celltype and sample independently
```{r}
autoscale <- FALSE
folder <- "pat_specific/" # "whole_dataset/" or "pat_specific/"
img_filename <- "_vis55_2000varfeatures.jpg"
plot_individual_maps <- function(spls, assay_name, folder, img_filename, autoscale = FALSE, patient_specific = TRUE, wd = ".") {
    features <- rownames(spls[[1]][[assay_name]])
    samples <- names(spls)
    for (sample in samples) {
        vis <- spls[[sample]]
        DefaultAssay(vis) <- assay_name
        subfolder <- if (autoscale) "autoscaled" else "scale_0_to_1"
        # Choose output folder based on patient_specific
        outdir <- paste0(wd, "map_decon/individual_maps/", folder, sample, "/", subfolder)
        if (!dir.exists(outdir)) {
            dir.create(outdir, recursive = TRUE)
        }
        if (length(vis$sample_id) > 5000) {
            pt_size <- 0.7
        } else if (length(vis$sample_id) < 5000) {
            pt_size <- 1.2
        }
        for (feature in features){
            p <- MapFeatures(vis, features = feature, colors = viridis::turbo(11), label_by = "sample_id", pt_size = pt_size, override_plot_dims = TRUE, scale = "shared", ncol = 1)
            if (!autoscale) {
                p <- p & scale_fill_viridis_c(limits = c(0,1), option = "H") # make scale 0 to 1
            }
            ggsave(
                filename = paste0(outdir, "/", gsub("/", "_", feature), img_filename),
                plot = p,
                height = 4,
                width = 4,
                dpi = 600,
                limitsize = FALSE
            )
        }
    }
}
```

##### merge Visium object
```{r}
if (length(spls) > 1) {
    mergedVis <- MergeSTData(spls[[1]], spls[-1])
} else {
    mergedVis <- spls[[1]]
}
mergedVis <- JoinLayers(mergedVis)
table(mergedVis$sample_id)
# saveRDS(mergedVis, paste0(rootObj, "merged_visium_deconSubpopulations_simplex.rds"))
```

```{r}
spls <- list()
samples <- unique(mergedVis$sample_id)
for (sample in samples){
    spl <- SubsetSTData(mergedVis, expression = sample_id == !!sample)
    spls[[sample]] <- spl
}
```

##### Spatial correlation

###### defining colors for each type
```{r}
# Define categories based on patient or subtype information
categories <- list(
    "apCAF_classical" = c("shared", "stromal"),
    "apCAF_TcellMimicry" = c("shared", "stromal"),
    "iCAF_classical" = c("shared", "stromal"),
    "iCAF_secretory" = c("shared", "stromal"),
    "iCAF_lipid-associated" = c("shared", "stromal"),
    "mCAF_plastic" = c("shared", "stromal"),
    "mCAF_classical" = c("shared", "stromal"),
    "mCAF_remodeling" = c("shared", "stromal"),
    "ifnCAF_interferon" = c("shared", "stromal"),
    "mCAF_patient4-specific" = c("patient4", "stromal"),
    "vCAF" = c("shared", "stromal"),
    "Adipocytes" = c("shared", "stromal"),
    "PVL" = c("shared", "stromal"),
    "Endothelial" = c("shared", "stromal"),
    "LumProgenitors_shared" = c("shared", "epithelial"),
    "Myoepithelial" = c("shared", "epithelial"),
    "LumProgenitors_transitional_pat4" = c("patient4", "epithelial"),
    "LumProgenitors_Cycl_pat4" = c("patient4", "epithelial"),
    "Cancer_Basal/Cycl_pat4" = c("patient4", "epithelial"),
    "Cancer_Basal/Plastic_pat4" = c("patient4", "epithelial"),
    "Cancer_Transitional_pat4" = c("patient4", "epithelial"),
    "Cancer_Cycl_pat2" = c("patient2", "epithelial"),
    "Cancer_LumB_pat2" = c("patient2", "epithelial"),
    "Cancer_Basal/LumB_pat2" = c("patient2", "epithelial"),
    "Cancer_Basal_pat9" = c("patient9", "epithelial"),
    "Cancer_Basal/LumA_pat1" = c("patient1", "epithelial"),
    "Cancer_Lum_pat7" = c("patient7", "epithelial"),
    "Cancer_LumA_pat1" = c("patient1", "epithelial"),
    "Cancer_LumA_pat8" = c("patient8", "epithelial"),
    "Cancer_Her2_pat9" = c("patient9", "epithelial"),
    "Cancer_Basal_stem-like_pat6" = c("patient6", "epithelial"),
    "Cancer_Basal_stress_pat6" = c("patient6", "epithelial"),
    "Cancer_Plastic_pat9" = c("patient9", "epithelial"),
    "Cancer_Basal_EMT_pat5" = c("patient5", "epithelial"),
    "Cancer_Basal_AntigPres_pat5" = c("patient5", "epithelial"),
    "Cancer_Basal_metabolic_pat5" = c("patient5", "epithelial"),
    "T-cells_atypical" = c("shared", "immune"),
    "T-cells_CD8_effector" = c("shared", "immune"),
    "T-cells_CD4_naive" = c("shared", "immune"),
    "T-cells_CD4_Treg" = c("shared", "immune"),
    "B-cells_activated" = c("shared", "immune"),
    "B/T-mixed_patient4" = c("patient4", "immune"),
    "immune_undefined" = c("shared", "immune"),
    "T-cells_cycling" = c("shared", "immune"),
    "Myeloid_atypical_patient6" = c("patient6", "immune"),
    "cDC2-like" = c("shared", "immune"),
    "cDC1-like" = c("shared", "immune"),
    "Neutrophils" = c("shared", "immune"),
    "Macrophages_monocyte_derived" = c("shared", "immune"),
    "pDCs" = c("shared", "immune"),
    "Mast_cells" = c("shared", "immune"),
    "Macrophages_inflammatory" = c("shared", "immune"),
    "Macrophages_interferon" = c("shared", "immune"),
    "Macrophages_immunoregulatory" = c("shared", "immune"),
    "Macrophages_tissue_resident" = c("shared", "immune"),
    "Plasma_cells" = c("shared", "immune")
)

celltypes_collapsed <- c(
  # --- Stromal (unchanged) ---
  "apCAF_classical" = "apCAF_classical",
  "apCAF_TcellMimicry" = "apCAF_TcellMimicry",
  "iCAF_classical" = "iCAF_classical",
  "iCAF_secretory" = "iCAF_secretory",
  "iCAF_lipid-associated" = "iCAF_lipid-associated",
  "mCAF_plastic" = "mCAF_plastic",
  "mCAF_classical" = "mCAF_classical",
  "mCAF_remodeling" = "mCAF_remodeling",
  "ifnCAF_interferon" = "ifnCAF_interferon",
  "mCAF_patient4-specific" = "mCAF_patient4-specific",
  "vCAF" = "vCAF",
  "Adipocytes" = "Adipocytes",
  "PVL" = "PVL",
  "Endothelial" = "Endothelial",

  # --- Epithelial (collapsed) ---
  "Cancer_Basal/Cycl_pat4" = "Cancer_Basal",
  "Cancer_Basal/Plastic_pat4" = "Cancer_Basal",
  "Cancer_Transitional_pat4" = "Cancer_Transitional",
  "Cancer_Cycl_pat2" = "Cancer_Cycling",
  "Cancer_LumB_pat2" = "Cancer_LumB",
  "Cancer_Basal/LumB_pat2" = "Cancer_LumB",
  "Cancer_Basal_pat9" = "Cancer_Basal",
  "Cancer_Basal/LumA_pat1" = "Cancer_LumA",
  "Cancer_Lum_pat7" = "Cancer_Lum",
  "Cancer_LumA_pat1" = "Cancer_LumA",
  "Cancer_LumA_pat8" = "Cancer_LumA",
  "Cancer_Her2_pat9" = "Cancer_Her2",
  "Cancer_Basal_stem-like_pat6" = "Cancer_Basal",
  "Cancer_Basal_stress_pat6" = "Cancer_Basal",
  "Cancer_Plastic_pat9" = "Cancer_Other",
  "Cancer_Basal_EMT_pat5" = "Cancer_Basal",
  "Cancer_Basal_AntigPres_pat5" = "Cancer_Basal",
  "Cancer_Basal_metabolic_pat5" = "Cancer_Basal",
  "LumProgenitors_shared" = "LumProgenitors",
  "Myoepithelial" = "Myoepithelial",
  "LumProgenitors_transitional_pat4" = "LumProgenitors_transitional",
  "LumProgenitors_Cycl_pat4" = "LumProgenitors",

  # --- Immune (unchanged) ---
  "T-cells_atypical" = "T-cells_atypical",
  "T-cells_CD8_effector" = "T-cells_CD8_effector",
  "T-cells_CD4_naive" = "T-cells_CD4_naive",
  "T-cells_CD4_Treg" = "T-cells_CD4_Treg",
  "B-cells_activated" = "B-cells_activated",
  "B/T-mixed_patient4" = "B/T-mixed_patient4",
  "immune_undefined" = "immune_undefined",
  "T-cells_cycling" = "T-cells_cycling",
  "Myeloid_atypical_patient6" = "Myeloid_atypical_patient6",
  "cDC2-like" = "cDC2-like",
  "cDC1-like" = "cDC1-like",
  "Neutrophils" = "Neutrophils",
  "Macrophages_monocyte_derived" = "Macrophages_monocyte_derived",
  "pDCs" = "pDCs",
  "Mast_cells" = "Mast_cells",
  "Macrophages_inflammatory" = "Macrophages_inflammatory",
  "Macrophages_interferon" = "Macrophages_interferon",
  "Macrophages_immunoregulatory" = "Macrophages_immunoregulatory",
  "Macrophages_tissue_resident" = "Macrophages_tissue_resident",
  "Plasma_cells" = "Plasma_cells"
)
# Create a mapping from collapsed names to specificity and celltype
categories_collapsed <- list(
    # Stromal
    "apCAF_classical" = c("NA", "stromal"),
    "apCAF_TcellMimicry" = c("NA", "stromal"),
    "iCAF_classical" = c("NA", "stromal"),
    "iCAF_secretory" = c("NA", "stromal"),
    "iCAF_lipid-associated" = c("NA", "stromal"),
    "mCAF_plastic" = c("NA", "stromal"),
    "mCAF_classical" = c("NA", "stromal"),
    "mCAF_remodeling" = c("NA", "stromal"),
    "ifnCAF_interferon" = c("NA", "stromal"),
    "mCAF_patient4-specific" = c("NA", "stromal"),
    "vCAF" = c("NA", "stromal"),
    "Adipocytes" = c("NA", "stromal"),
    "PVL" = c("NA", "stromal"),
    "Endothelial" = c("NA", "stromal"),

    # Epithelial
    "Cancer_Basal" = c("NA", "epithelial"),
    "Cancer_Transitional" = c("NA", "epithelial"),
    "Cancer_Cycling" = c("NA", "epithelial"),
    "Cancer_LumB" = c("NA", "epithelial"),
    "Cancer_LumA" = c("NA", "epithelial"),
    "Cancer_Lum" = c("NA", "epithelial"),
    "Cancer_Her2" = c("NA", "epithelial"),
    "Cancer_Other" = c("NA", "epithelial"),
    "LumProgenitors" = c("NA", "epithelial"),
    "LumProgenitors_transitional" = c("NA", "epithelial"),
    "Myoepithelial" = c("NA", "epithelial"),

    # Immune
    "T-cells_atypical" = c("NA", "immune"),
    "T-cells_CD8_effector" = c("NA", "immune"),
    "T-cells_CD4_naive" = c("NA", "immune"),
    "T-cells_CD4_Treg" = c("NA", "immune"),
    "B-cells_activated" = c("NA", "immune"),
    "B/T-mixed_patient4" = c("NA", "immune"),
    "immune_undefined" = c("NA", "immune"),
    "T-cells_cycling" = c("NA", "immune"),
    "Myeloid_atypical_patient6" = c("NA", "immune"),
    "cDC2-like" = c("NA", "immune"),
    "cDC1-like" = c("NA", "immune"),
    "Neutrophils" = c("NA", "immune"),
    "Macrophages_monocyte_derived" = c("NA", "immune"),
    "pDCs" = c("NA", "immune"),
    "Mast_cells" = c("NA", "immune"),
    "Macrophages_inflammatory" = c("NA", "immune"),
    "Macrophages_interferon" = c("NA", "immune"),
    "Macrophages_immunoregulatory" = c("NA", "immune"),
    "Macrophages_tissue_resident" = c("NA", "immune"),
    "Plasma_cells" = c("NA", "immune")
)
# Convert underscores to dashes in all collapsed category names
names(categories_collapsed) <- gsub("_", "-", names(categories_collapsed))

# Convert underscores to dashes in all category names
names(categories) <- gsub("_", "-", names(categories))
names(categories_collapsed) <- gsub("_", "-", names(categories_collapsed))

# Define corresponding colors for each category
category_colors <- list(
    specificity = c(
        "shared" = "#27AE60",
        "patient1" = "#F1C40F",
        "patient2" = "#1ABC9C",
        "patient4" = "#E67E22",
        "patient5" = "#F39C12",
        "patient6" = "#C0392B",
        "patient7" = "#16A085",
        "patient8" = "#34495E",
        "patient9" = "#6C3483"
    )
    ,
    celltype = c(
        "immune" = "#3498DB",
        "stromal" = "#2ECC71",
        "epithelial" = "#E74C3C"
    )
)
# Define colors for collapsed categories (celltype and specificity)
category_colors_collapsed <- list(
    specificity = c(
        "NA" = "NA"
    ),
    celltype = c(
        "immune" = "#3498DB",
        "stromal" = "#2ECC71",
        "epithelial" = "#E74C3C"
    )
)
```

###### per patient 
```{r}
for (spl in spls){
    DefaultAssay(spl) <- "subpopulation"
    data <- FetchData(spl, rownames(spl)) %>% 
        as.data.frame() %>%
        select(where(~ sum(., na.rm = TRUE) > 0))  # Add na.rm=TRUE to handle NAs

    cor_matrix <- cor(data, method = "spearman", use = "pairwise.complete.obs")
    diag(cor_matrix) <- NA

    valid_rows <- intersect(rownames(cor_matrix), names(categories))
    specificity_pt <- do.call(rbind, lapply(categories[valid_rows], function(x) {
        c(category_colors$specificity[x[1]], category_colors$celltype[x[2]])
    }))

    # Generate heatmap and save as PDF
    pdf(paste0(wd, "correlation/", unique(spl$sample_id), "_correlation_spatial_finalSubpopulations.pdf"), width = 11, height = 8)
    heatmap3(
        cor_matrix, 
        margins = c(16, 16), 
        symm = TRUE,
        RowSideColors = specificity_pt[, 2:1],  
        ColSideColors = specificity_pt,
        col = colorRampPalette(c("blue", "white", "red"))(20), 
        balanceColor = TRUE, 
        RowSideLabs = c("CellType", "Specificity"),  # Label for RowSideColors
        ColSideLabs = c("Specificity", "CellType")  # Label for ColSideColors
    )

    legend(
    "topright",  
    legend = names(category_colors$specificity),  
    fill = category_colors$specificity,  
    title = "Specificity",
    inset = c(0, 0)
    )
    legend(
    "topright",  
    legend = names(category_colors$celltype),  
    fill = category_colors$celltype,  
    title = "CellType",
    inset = c(0, 0.4)
    )
    dev.off()
    # Save correlation matrix as CSV
    write.csv(cor_matrix, file = paste0(wd, "correlation/", unique(spl$sample_id), "_correlation_matrix.csv"), row.names = TRUE)
}

```

### collapse epithelial to broad categories 
In order to study correlations of shared signatures across patients, we collapse the epithelial categories to broader categories
in order to generate more robust correlations. 
```{r}
# Collapse subpopulation proportions according to celltypes_collapsed mapping
subpop_mat <- GetAssayData(mergedVis, assay = "subpopulation", slot = "data")
collapsed_names <- celltypes_collapsed[rownames(subpop_mat)]
collapsed_names <- as.character(collapsed_names)
# Remove any NA mappings
valid_idx <- !is.na(collapsed_names)
subpop_mat <- subpop_mat[valid_idx, , drop = FALSE]
collapsed_names <- collapsed_names[valid_idx]

# Aggregate (sum) rows by collapsed category
collapsed_mat <- as.matrix(rowsum(as.matrix(subpop_mat), group = collapsed_names))
# Ensure order is consistent and unique
collapsed_mat <- collapsed_mat[unique(collapsed_names), , drop = FALSE]

# Add as new assay
mergedVis[["subpopulation_collapsed"]] <- CreateAssayObject(data = collapsed_mat)

# split 
spls <- list()
samples <- unique(mergedVis$sample_id)
for (sample in samples){
    spl <- SubsetSTData(mergedVis, expression = sample_id == !!sample)
    spls[[sample]] <- spl
}

# plot 
plot_individual_maps(spls, assay_name = "subpopulation_collapsed", folder = "collapsed_epithelial/", img_filename = "_vis55_2000varfeatures_collapsed.jpg", autoscale = FALSE, patient_specific = patient_specific, wd = wd)
```

##### Subtype specific 
```{r}
# Create a vector with matches between sample and subtype
patient_subtypes <- c(
    "patient1_55um" = "LuminalA",
    "patient2_55um" = "LuminalB(ER)",
    "patient3_55um" = "TNBC",
    "patient4_55um" = "TNBC",
    "patient4_HD" = "TNBC",
    "patient5_55um" = "TNBC",
    "patient5_HD" = "TNBC",
    "patient6_55um" = "TNBC",
    "patient7_55um" = "LuminalB(ER)",
    "patient8_55um" = "LuminalA",
    "patient9_55um" = "LuminalB(ER)"
)
sbtps <- list()
subtypes <- unique(patient_subtypes)
for (subtype in subtypes){
    sbt <- SubsetSTData(mergedVis, expression = sample_id %in% names(patient_subtypes[patient_subtypes == subtype]))
    sbtps[[subtype]] <- sbt
}
```

```{r}
for (subtype in subtypes){
    sbt <- sbtps[[subtype]]
    DefaultAssay(sbt) <- "subpopulation_collapsed"
    # Fetch data and compute correlation
    data <- FetchData(sbt, rownames(sbt)) %>% 
        as.data.frame() %>%
        select(where(~ sum(., na.rm = TRUE) > 0))  # Add na.rm=TRUE to handle NAs
    cor_matrix <- cor(data, method = "spearman", use = "pairwise.complete.obs")
    diag(cor_matrix) <- NA

    valid_rows <- intersect(rownames(cor_matrix), names(categories_collapsed))
    specificity_pt <- do.call(rbind, lapply(categories_collapsed[valid_rows], function(x) {
        c(category_colors_collapsed$specificity[x[1]], category_colors_collapsed$celltype[x[2]])
    }))

    # Adapt figure size to number of valid_rows (smaller)
    n_valid <- length(valid_rows)
    print(n_valid)
    fig_width <- max(4, min(0.22 * n_valid, 15))
    fig_height <- max(4, min(0.22 * n_valid, 15))

    # Generate heatmap and save as PDF
    pdf(paste0(wd, "correlation/", subtype, "_correlation_finalSubpopulations.pdf"), width = fig_width, height = fig_height)
    heatmap3(
        cor_matrix, 
        margins = c(16, 16), 
        symm = TRUE,
        RowSideColors = specificity_pt[, 2],  
        ColSideColors = specificity_pt[, 2],
        col = colorRampPalette(c("blue", "white", "red"))(100),
        balanceColor = TRUE, 
        RowSideLabs = c("CellType"),  # Label for RowSideColors
        ColSideLabs = c("CellType")  # Label for ColSideColors
    )

    legend(
        "topright",  
        legend = names(category_colors_collapsed$celltype),  
        fill = category_colors_collapsed$celltype,  
        title = "CellType",
        x = "topright",
        inset = c(0, 0),
        bty = "n",  # Remove the box from the whole legend
        cex = 0.6   # Reduce legend text size
    )

    dev.off()
    # Save correlation matrix as CSV
    write.csv(cor_matrix, file = paste0(wd, "correlation/", subtype, "_correlation_matrix.csv"), row.names = TRUE)
}
```

##### merged
```{r}
DefaultAssay(mergedVis) <- "subpopulation_collapsed"

# Fetch data and compute correlation
data <- FetchData(mergedVis, rownames(mergedVis)) %>% 
    as.data.frame() %>%
    select(where(~ sum(., na.rm = TRUE) > 0))  # Add na.rm=TRUE to handle NAs
cor_matrix <- cor(data, method = "spearman", use = "pairwise.complete.obs")
diag(cor_matrix) <- NA

    valid_rows <- intersect(rownames(cor_matrix), names(categories_collapsed))
    specificity_pt <- do.call(rbind, lapply(categories_collapsed[valid_rows], function(x) {
        c(category_colors_collapsed$specificity[x[1]], category_colors_collapsed$celltype[x[2]])
    }))

# Generate heatmap and save as PDF
pdf(paste0(wd, "correlation/correlation_finalSubpopulations.pdf"), width = 12, height = 8)
heatmap3(
    cor_matrix, 
    margins = c(12, 12), 
    symm = TRUE,
    RowSideColors = specificity_pt[, 2],  
    ColSideColors = specificity_pt[, 2],
    col = colorRampPalette(c("blue", "white", "red"))(100),
    balanceColor = TRUE, 
    RowSideLabs = c("CellType"),  # Label for RowSideColors
    ColSideLabs = c("CellType")  # Label for ColSideColors
)

legend(
    "topright",  
    legend = names(category_colors_collapsed$celltype),  
    fill = category_colors_collapsed$celltype,  
    title = "CellType",
    x = "topright",
    inset = c(0, 0),
    bty = "n",  # Remove the box from the whole legend
    cex = 1   # Reduce legend text size
)

dev.off()
# Save correlation matrix as CSV
write.csv(cor_matrix, file = paste0(wd, "correlation/correlation_matrix.csv"), row.names = TRUE)
```

## marker genes

```{r}
Idents(int.all) <- "subpopulation"

markers <- FindAllMarkers(int.all, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)
top_markers <- markers %>%
    group_by(cluster) %>%
    top_n(n = 2, wt = avg_log2FC)
top100 <- markers %>%
    group_by(cluster) %>%
    top_n(n = 100, wt = avg_log2FC)
write.csv(top100, file = paste0(wd, "marker_genes/subpopulation_markers_top100.csv"), row.names = FALSE)

p <- DotPlot(int.all, features = unique(top_markers$gene)) + 
                coord_flip() +
                scale_color_gradientn(colors = c("blue", "white", "darkred")) +
                theme(axis.text.x = element_text(angle = 65, hjust = 1, vjust = 1))
ggsave(paste0(wd, "marker_genes/subpopulation_markers_dotplot.pdf"), p, width = 15, height = 20)
```

## GSEA

```{r}
library(singleseqgset)
library(msigdbr)

#### Select Gene Set of interest
# H: hallmark gene sets
# C1: positional gene sets
# C2: curated gene sets

category <- "H" # category from Human MSigDB Collections
h.human <- msigdbr(species="Homo sapiens",category=category)
subcategory <- ""
h.human <- h.human %>% filter(gs_subcat == subcategory)
h.names <- unique(h.human$gs_name)
h.sets <- vector("list",length=length(h.names))
names(h.sets) <- h.names
for (i in names(h.sets)) {
    h.sets[[i]] <- pull(h.human[h.human$gs_name==i,"gene_symbol"])
}
#### Filter out non relevant pathways
# Define the list of pathways to remove
remove_pathways <- c(
  "HALLMARK_BILE_ACID_METABOLISM",         # Liver-specific  
  "HALLMARK_PANCREAS_BETA_CELLS",          # Pancreatic function  
  "HALLMARK_SPERMATOGENESIS",              # Testis-specific  
  "HALLMARK_MYOGENESIS",                   # Muscle development  
  "HALLMARK_ADIPOGENESIS",                 # Fat cell differentiation  
  "HALLMARK_FATTY_ACID_METABOLISM",        # General lipid metabolism  
  "HALLMARK_HEME_METABOLISM",              # Blood-related, not CAF-specific  
  "HALLMARK_CHOLESTEROL_HOMEOSTASIS",      # Unrelated lipid metabolism  
  "HALLMARK_PEROXISOME",                   # General cell detox pathway  
  "HALLMARK_XENOBIOTIC_METABOLISM",        # Detox, not breast cancer-related  
  "HALLMARK_UV_RESPONSE_DN",               # Skin cancer-related  
  "HALLMARK_UV_RESPONSE_UP"                # Skin cancer-related  
)
# Filter the list, keeping only relevant pathways
h.sets <- h.sets[!names(h.sets) %in% remove_pathways]
##### Change names (aesthetic purposes only)
names(h.sets) <- gsub("HALLMARK_", "", names(h.sets))
#### Log Fold change
Idents(int.all) <- "subpopulation"
int.all$seurat_clusters <- Idents(int.all)
mtx <- GetAssayData(int.all, assay = "RNA", layer = "data")
logfc <- logFC(cluster.ids = int.all@meta.data$seurat_clusters, expr.mat = mtx)
names(logfc)
#### Run GSEA
gse.res <- wmw_gsea(expr.mat=mtx, cluster.cells=logfc[[1]],
                    log.fc.cluster=logfc[[2]],
                    gene.sets=h.sets)
res.stats <- gse.res[["GSEA_statistics"]]
res.pvals <- gse.res[["GSEA_p_values"]]
res.pvals <- apply(res.pvals,2,p.adjust,method="fdr") #Correct for multiple comparisons
res.stats[order(res.stats[,1],decreasing=TRUE)[1:10],] #Top gene sets enriched by z scores
#Plot the z scores with heatmap3
pdf(paste0(wd, "gsea/subpopulations_GSEA.pdf"), width = 15, height = 10)
heatmap3(res.stats, margins = c(10, 10))
dev.off()
```

### Markers and UMAP for major lineages
```{r}
# Separate into each lineage and plot UMAP and marker genes

# Use the 'categories' list to assign lineage for each subpopulation
subpop_names <- as.character(int.all$subpopulation)
subpop_names_dash <- gsub("_", "-", subpop_names) # replace underscores with dashes
cell_names <- colnames(int.all)
lineage_map <- sapply(subpop_names_dash, function(x) {
    if (!is.null(categories[[x]])) categories[[x]][2] else NA
})
names(lineage_map) <- cell_names
int.all$lineage <- lineage_map

# Define major lineages present in the data
lineages <- unique(na.omit(int.all$lineage))

# For each lineage: subset, plot UMAP, and marker genes
for (lin in lineages) {
        obj <- subset(int.all, subset = lineage == lin)
        if (ncol(obj) < 10) next  # skip if too few cells

        # UMAP colored by subpopulation
        p_umap <- DimPlot(obj, reduction = "umap", group.by = "subpopulation", label = TRUE) +
                ggtitle(paste("UMAP -", lin))

        # Find top marker genes for each subpopulation
        Idents(obj) <- "subpopulation"
        markers <- FindAllMarkers(obj, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)
        top_markers <- markers %>%
                group_by(cluster) %>%
                top_n(n = 3, wt = avg_log2FC)
        marker_genes <- unique(top_markers$gene)

        # DotPlot for marker genes
        p_dot <- DotPlot(obj, features = marker_genes) +
                coord_flip() +
                scale_color_gradientn(colors = c("blue", "white", "darkred")) +
                theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
                ggtitle(paste("Markers -", lin))

        # Save plots
        # ggsave(filename = paste0(wd, "umaps/umap_", lin, ".pdf"), plot = p_umap, width = 8, height = 6)
        ggsave(filename = paste0(wd, "marker_genes/markers_dotplot_", lin, ".pdf"), plot = p_dot, width = 10, height = 8)
}
```

#### manually plot umaps 
```{r}
spl <- readRDS(paste0(DATA_ROOT, "/r_objects/breast_cancer/single_nuclei/cross_patient/caf_subpop_analysis.rds"))
Idents(spl) <- "seurat_clusters"
populations <- names(categories)[sapply(categories, function(x) x[2] == "caf")]
# Ensure populations match the Idents, accounting for underscores/dashes
idents_levels <- levels(Idents(spl))
populations <- gsub("_", "-", populations)
idents_levels <- gsub("_", "-", idents_levels)
populations <- intersect(populations, idents_levels)
# Ensure both Idents and populations use dashes
Idents(spl) <- gsub("_", "-", Idents(spl))
populations <- gsub("_", "-", populations)
spl <- subset(spl, idents = populations)
ggsave(
    plot = DimPlot(spl, reduction = "umap", group.by = "seurat_clusters", label = FALSE, repel = TRUE)
     + theme(legend.position = "none"),
    filename = paste0(wd, "umaps/umap_CAF_subpopulations_nolegend.pdf"),
    width = 4,
    height = 4
)
```

## Distribution of Patients, subtypes across Subpopulations
```{r}
entities <- "subpopulation"
p1 <- plot_distribution(int.all, entities, "patient_ID")
p2 <- plot_distribution(int.all, entities, "subtype")
p <- p1 + p2

ggsave(
    filename = paste0(wd, "distribution/", entities, "_distribution.pdf"),
    plot = p,
    width = 20,
    height = 4.5
)
```

# Network visualization of subpopulation compartments
```{r}
library(igraph)
wd <- paste0(FIGS_ROOT, "/breast_cancer/analysis_cellStateLevel/final_subpopulations/")
int.all <- readRDS(paste0(rootObj, "integrated_all_subpopulations.rds"))
mergedVis <- readRDS(paste0(rootObj, "merged_visium_deconSubpopulations_simplex.rds"))
spls <- list()
samples <- unique(mergedVis$sample_id)
for (sample in samples){
    spl <- SubsetSTData(mergedVis, expression = sample_id == !!sample)
    spls[[sample]] <- spl
}
# Create a vector with matches between sample and subtype
patient_subtypes <- c(
    "patient1_55um" = "LuminalA",
    "patient2_55um" = "LuminalB(ER)",
    "patient3_55um" = "TNBC",
    "patient4_55um" = "TNBC",
    "patient4_HD" = "TNBC",
    "patient5_55um" = "TNBC",
    "patient5_HD" = "TNBC",
    "patient6_55um" = "TNBC",
    "patient7_55um" = "LuminalB(ER)",
    "patient8_55um" = "LuminalA",
    "patient9_55um" = "LuminalB(ER)"
)
sbtps <- list()
subtypes <- unique(patient_subtypes)
for (subtype in subtypes){
    sbt <- SubsetSTData(mergedVis, expression = sample_id %in% names(patient_subtypes[patient_subtypes == subtype]))
    sbtps[[subtype]] <- sbt
}
```

### generate spatial correlation network
#### merged
```{r}
set.seed(1)
corMat_network <- read.csv(paste0(wd, "correlation/correlation_matrix.csv"), row.names = 1) |> t()
corMat_network[corMat_network<0.2] <- 0
network <- graph_from_adjacency_matrix(corMat_network, weighted=T, mode="undirected", diag=F)

# Count the number of degree for each node:
deg <- degree(network, mode="all")

# Assign colors based on celltype
node_names <- V(network)$name
celltype_colors <- sapply(node_names, function(n) {
    ct <- categories_collapsed[[n]][2]  # 2nd element is "celltype"
    if (!is.null(ct) && ct %in% names(category_colors_collapsed$celltype)) {
        return(category_colors_collapsed$celltype[[ct]])
    } else {
        return("#BDC3C7")  # fallback color (light grey)
    }
})

# Plot
layout <- layout_with_fr(network)
pdf(paste0(wd, "niches/graph/network_visualization.pdf"), width = 5, height = 5)
par(mar = c(2, 2, 2, 2))  # Broaden right margin
plot(network,
    layout = layout,
    vertex.size = pmax(deg * 2.7, 4),  # minimum size for visibility
    vertex.color = celltype_colors,
    vertex.frame.color = "white",
    vertex.label.cex = 0.7,
    vertex.label.family = "Helvetica",
    vertex.label.color = "black",
    edge.width = pmin(E(network)$weight * 3, 5),  # cap edge width
    edge.color = rgb(0.6, 0.6, 0.6, 0.3),  # light grey transparent edges
    main = "All Samples")
# Add legend for celltype colors
# legend(
#     "topright",  
#     legend = names(category_colors$celltype),  
#     fill = category_colors$celltype,  
#     title = "CellType",
#     inset = c(0, 0)  # Adjust inset further to move it more to the right
# )
dev.off()
```

#### per patient
Here we dont used the collapsed categories since we want to see the patient specific populations
```{r}
sampleNetworks <- list()
for (spl in spls){
    DefaultAssay(spl) <- "subpopulation"
    sample <- unique(spl$sample_id)
    data <- FetchData(spl, rownames(spl)) %>% 
        as.data.frame() %>%
        select(where(~ sum(., na.rm = TRUE) > 0))  # Add na.rm=TRUE to handle NAs

    cor_matrix <- cor(data, method = "spearman", use = "pairwise.complete.obs")
    diag(cor_matrix) <- NA

    set.seed(1)
    corMat_network <- cor_matrix |> t()
    corMat_network[corMat_network<0.15] <- 0
    network <- graph_from_adjacency_matrix(corMat_network, weighted=T, mode="undirected", diag=F)

    # Count the number of degree for each node:
    deg <- degree(network, mode="all")
    
    # Assign colors based on celltype
    node_names <- V(network)$name
    celltype_colors <- sapply(node_names, function(n) {
        ct <- categories[[n]][2]  # 2nd element is "celltype"
        if (!is.null(ct) && ct %in% names(category_colors$celltype)) {
            return(category_colors$celltype[[ct]])
        } else {
            return("#BDC3C7")  # fallback color (light grey)
        }
    })
    
    # Plot
    layout <- layout_with_fr(network)
    pdf(paste0(wd, "niches/graph/network_", sample, ".pdf"), width = 5, height = 5)
    par(mar = c(2, 2, 2, 2))  # Broaden right margin
    plot(network,
        layout = layout,
        vertex.size = pmax(deg * 2.7, 4),  # minimum size for visibility
        vertex.color = celltype_colors,
        vertex.frame.color = "white",
        vertex.label.cex = 0.7,
        vertex.label.family = "Helvetica",
        vertex.label.color = "black",
        edge.width = pmin(E(network)$weight * 3, 5),  # cap edge width
        edge.color = rgb(0.6, 0.6, 0.6, 0.3),  # light grey transparent edges
        main = paste(sample))
    # Add legend for celltype colors
    # legend(
    #     "topright",  
    #     legend = names(category_colors$celltype),  
    #     fill = category_colors$celltype,  
    #     title = "CellType",
    #     inset = c(0, 0)  # Adjust inset further to ensure it fits within the plot area
    # )
    dev.off()

    sampleNetworks[[sample]] <- network
}
```

#### per subtype
```{r}
subtypeNetworks <- list()
for (subtype in subtypes){
    set.seed(1)
    corMat_network <- read.csv(paste0(wd, "correlation/",subtype, "_correlation_matrix.csv"), row.names = 1) |> t()
    corMat_network[corMat_network<0.2] <- 0
    network <- graph_from_adjacency_matrix(corMat_network, weighted=T, mode="undirected", diag=F)

    # Count the number of degree for each node:
    deg <- degree(network, mode="all")

    # Assign colors based on celltype
    node_names <- V(network)$name
    celltype_colors <- sapply(node_names, function(n) {
        ct <- categories_collapsed[[n]][2]  # 2nd element is "celltype"
        if (!is.null(ct) && ct %in% names(category_colors_collapsed$celltype)) {
            return(category_colors_collapsed$celltype[[ct]])
        } else {
            return("#BDC3C7")  # fallback color (light grey)
        }
    })

    subtypeNetworks[[subtype]] <- network
    
    # Plot
    layout <- layout_with_fr(network)
    pdf(paste0(wd, "niches/graph/network_", subtype, ".pdf"), width = 5, height = 5)
    par(mar = c(2, 2, 2, 2))  # Broaden right margin
    plot(network,
        layout = layout,
        vertex.size = pmax(deg * 2.7, 4),  # minimum size for visibility
        vertex.color = celltype_colors,
        vertex.frame.color = "white",
        vertex.label.cex = 0.7,
        vertex.label.family = "Helvetica",
        vertex.label.color = "black",
        edge.width = pmin(E(network)$weight * 3, 5),  # cap edge width
        edge.color = rgb(0.6, 0.6, 0.6, 0.3),  # light grey transparent edges
        main = paste(subtype))
    # Add legend for celltype colors
    # legend(
    #     "topright",  
    #     legend = names(category_colors$celltype),  
    #     fill = category_colors$celltype,  
    #     title = "CellType",
    #     inset = c(0, 0)  # Adjust inset further to ensure it fits within the plot area
    # )
    dev.off()
}
```

### Cluster based on graph connectivity - louvain algorithm
```{r}
recolored <- TRUE # we go back to these plots after recoloring in downstream analysis 
cluster <- FALSE
networks <- sampleNetworks  # Choose which networks to cluster
# Function to get clusters for each subtype network
if (cluster) {
    get_clusters <- function(network) {
    cl <- cluster_louvain(network, resolution = 1.7)
    memb <- membership(cl)
    # memb[!(memb %in% which(sizes(cl) > 1))] <- NA
    memb
    }
    # Apply clustering to each subtype network
    sample_clusters <- lapply(networks, get_clusters)
}

# Add sample clusters to graph plot
for (sample in names(sample_clusters)) {
    network <- networks[[sample]]
    membership <- sample_clusters[[sample]]

        # Assign colors based on niche membership and subtype
        cluster_colors <- sapply(names(membership), function(n) {
            if (recolored) {
                niche_name <- paste0(sample, "-niche", membership[n])
                color_map <- niche_color_maps[[sample]]
                # Convert underscores to dashes in sample for matching
                sample_dash <- gsub("_", "-", sample)
                niche_name_dash <- paste0(sample_dash, "-niche", membership[n])
                if (!is.null(color_map) && niche_name_dash %in% names(color_map)) {
                    return(color_map[niche_name_dash])
                } else {
                    return("#BDC3C7")  # fallback color (light grey)
                }
            } else {
                rainbow(length(unique(membership)))[membership[n]]
            }
        })
    
    # Assign shapes based on cell type
    node_shapes <- sapply(names(membership), function(n) {
        celltype <- categories[[n]][2]  # Get cell type from categories
        if (!is.null(celltype) && celltype %in% c("immune", "stromal", "epithelial")) {
            switch(celltype,
                   "immune" = "circle",
                   "stromal" = "square",
                   "epithelial" = "sphere")
        } else {
            "circle"  # Default shape
        }
    })
    
    # Plot
    set.seed(1)  # Ensure reproducibility of layout
    layout <- layout_with_fr(network)
    if (recolored) {
        pdf(paste0(wd, "niches/graph/network_", sample, "_colorBy_niche.pdf"), width = 4, height = 4)
    } else {
        pdf(paste0(wd, "niches/graph/network_", sample, "_clusters.pdf"), width = 4, height = 4)
    }
    par(mar = c(2, 2, 2, 2))  # Broaden right margin
    plot(network,
        layout = layout,
        vertex.size = pmax(degree(network, mode="all") * 2.7, 4),  # minimum size for visibility
        vertex.color = cluster_colors,
        vertex.shape = node_shapes,
        vertex.frame.color = "white",
        vertex.label.cex = 0.5,
        vertex.label.family = "Helvetica",
        vertex.label.color = "black",
        edge.width = pmin(E(network)$weight * 3, 5),  # cap edge width
        edge.color = rgb(0.6, 0.6, 0.6, 0.3),  # light grey transparent edges
        main = paste(sample, "- Spatial Niches"))
    
    # # Add legend for clusters (color)
    # legend(
    #     "topright",  
    #     legend = if (recolored) {
    #         c(paste("Niche", unique(membership[cluster_colors != "#BDC3C7"])), "No niche")
    #     } else {
    #         paste("Cluster", unique(membership))
    #     },
    #     fill = if (recolored) {
    #         c(sapply(unique(membership[cluster_colors != "#BDC3C7"]), function(n) niche_colors[paste0(sample, "-niche", n)]), "#BDC3C7")
    #     } else {
    #         rainbow(length(unique(membership)))
    #     },
    #     title = if (recolored) "Niches" else "Clusters",
    #     inset = c(0, 0), 
    #     border = NA,
    #     cex = 0.3  # Make legend text smaller
    # )
    
    # # Add legend for shapes
    # legend(
    #     "bottomright",
    #     legend = c("Immune", "Stromal", "Epithelial"),
    #     pch = c(21, 22, 16),  # Corresponding shapes
    #     title = "Cell Lineage",
    #     inset = c(0, 0),
    #     col = "black",
    #     pt.bg = "white"
    # )
    dev.off()
}
```

## Create niche scores based on graph clusters
```{r}
niches <- list()
for (sample in names(sample_clusters)) {
    clusters <- sample_clusters[[sample]]
    unique_clusters <- unique(clusters)
    for (cluster_id in unique_clusters) {
        cluster_members <- names(clusters[clusters == cluster_id])
        if (length(cluster_members) > 1) {
            niche_name <- paste0(sample, "_niche", cluster_id)
            niches[[niche_name]] <- cluster_members
        }
    }
}
for (sample in names(sample_clusters)) {
    spl <- spls[[sample]]
    subpop_df <- as.data.frame(t(GetAssayData(spl, assay = "subpopulation", slot = "data")))
    niche_scores_list <- list()
    # Only use niches belonging to this sample
    sample_niches <- names(niches)[grepl(paste0("^", gsub("_", "-", sample), "-"), gsub("_", "-", names(niches)))]
    for (niche_name in sample_niches) {
        cell_types <- niches[[niche_name]]
        # Filter only the columns in subpop_df that match cell types in this niche
        common_cell_types <- intersect(cell_types, colnames(subpop_df))
        if (length(common_cell_types) > 0) {
            # Extract relevant columns
            niche_matrix <- subpop_df[, common_cell_types, drop = FALSE]
            # Compute row-wise mean of proportions
            niche_score <- rowSums(niche_matrix)
            # Store the vector
            niche_scores_list[[niche_name]] <- niche_score
        } else {
            warning(paste("No matching cell types found for", niche_name))
            # Fill with NA if no cell types matched
            niche_scores_list[[niche_name]] <- rep(NA, nrow(subpop_df))
        }
    }
    # Combine all niche scores into a data frame
    niche_score_mat <- as.data.frame(niche_scores_list, check.names = FALSE)
    rownames(niche_score_mat) <- rownames(subpop_df)
    # Transpose to get niches as rows (optional depending on downstream analysis)
    niche_score_mat <- t(as.matrix(niche_score_mat))
    # Z-score normalization across samples for each niche (row-wise)
    if (nrow(niche_score_mat) > 1) {
        niche_score_mat <- t(scale(t(niche_score_mat)))  # scales across columns (samples)
    }
    spl[["niche_score"]] <- CreateAssayObject(data = niche_score_mat)
    DefaultAssay(spl) <- "niche_score"
    spls[[sample]] <- spl
}
```

### map niche scores spatially
###### plot each niche score and sample independently
```{r}
autoscale <- TRUE
img_filename <- "_nicheScore.jpg"
folder <- "niches/map_score/"
DefaultAssay(mergedVis) <- "niche_score"
for (sample in samples) {
    vis <- spls[[sample]]
    features <- rownames(vis)
    subfolder <- if (autoscale) "autoscaled" else "same_scale"
    if (!dir.exists(paste0(wd, folder, sample, "/", subfolder))) {
        dir.create(paste0(wd, folder, sample, "/", subfolder), recursive = TRUE)
    }
        if (length(vis$sample_id) > 5000) {
        pt_size <- 0.7
    } else if (length(vis$sample_id) < 5000) {
        pt_size <- 1.2
    }
    for (feature in features){
            p <- MapFeatures(vis, features = feature, colors = viridis::turbo(11), label_by = "sample_id", pt_size = pt_size, override_plot_dims = TRUE, scale = "shared", ncol = 1)
            if (!autoscale) {
                p <- p & scale_fill_viridis_c(limits = c(-1.5,7), option = "H") # make scale n to n
            }
            if (!dir.exists(paste0(wd, folder, sample, "/", subfolder))) {
                dir.create(paste0(wd, folder, sample, "/", subfolder), recursive = TRUE)
            }
            ggsave(
                filename = paste0(wd, folder, sample, "/", subfolder, "/", gsub("/", "_", feature), img_filename),
                plot = p,
                height = 4,
                width = 4,
                dpi = 600,
                limitsize = FALSE
            )
    }
}
```

## Create discrete Niche labels in spatial data

###### Create maximum value metadata labels for the niche scores and plot labels 
```{r}
samples <- unique(mergedVis$sample_id)
niche_color_maps <- list()  # Store color maps for each sample

for (sample in samples) {
    vis <- spls[[sample]]

    niche_matrix <- GetAssayData(vis, assay = "niche_score", slot = "data")
    # Get the feature (row) with the max value per column (spot)
    max_features <- apply(niche_matrix, 2, function(x) {
        if (all(is.na(x))) {
            return(NA)  # No niche passes threshold
        } else {
            return(rownames(niche_matrix)[which.max(x)])
        }
    })
    vis$top_niche <- unlist(max_features)

    # Assign colors using rainbow palette based on unique top_niche values
    membership <- as.factor(vis$top_niche)
    niche_levels <- levels(membership)
    color_map <- setNames(rainbow(length(niche_levels)), niche_levels)
    colors <- color_map[as.character(vis$top_niche)]

    # Save color_map for this sample
    niche_color_maps[[sample]] <- color_map

    if (length(vis$sample_id) > 5000) {
        pt_size <- 1.4
    } else {
        pt_size <- 2.3
    }
    p <- MapLabels(
            vis,
            column_name = "top_niche",
            label_by = "sample_id",
            colors = color_map,
            pt_size = pt_size,
            override_plot_dims = TRUE,
            drop_na = TRUE
        ) &
        theme(legend.position = "right") &
        guides(fill = guide_legend(override.aes = list(size = 4)))
    ggsave(
        filename = paste0(wd, "niches/map_labels/", sample, "_niche_labels.jpg"),
        plot = p,
        height = 5,
        width = 7,
        dpi = 600,
        limitsize = FALSE
    )
}
```

### apply niche coloring scheme to patient-specific network plots. ### checkpoint 
we go back to previous code and set "recolored" to TRUE

## recoloring patient 4
```{r}
sample <- "patient4_55um"
network <- sampleNetworks[[sample]]

# Define your manual color mapping for patient4_55um
manual_colors_patient4 <- c(
    "patient4_55um-niche2" = "#3cab17ff",
    "patient4_55um-niche4" = "#98c9feff",
    "patient4_55um-niche5" = "#e9124fd3",
    "patient4_55um-niche8" = "#a146bfff",
    "patient4_55um-niche7" = "#eb8416ff"
    # Add or adjust as needed for your actual niche names
)

# Only plot for patient4_55um
sample <- "patient4_55um"
network <- networks[[sample]]
membership <- sample_clusters[[sample]]

# Assign colors based on manual mapping (fix dash/underscore issue)
cluster_colors <- sapply(names(membership), function(n) {
    niche_name <- paste0(sample, "-niche", membership[n])
    # Replace underscores with dashes for matching
    niche_name_dash <- gsub("_", "-", niche_name)
    manual_names_dash <- gsub("_", "-", names(manual_colors_patient4))
    color_idx <- match(niche_name_dash, manual_names_dash)
    if (!is.na(color_idx)) {
        manual_colors_patient4[[color_idx]]
    } else {
        "#BDC3C7"
    }
})

# Assign shapes based on cell type
node_shapes <- sapply(names(membership), function(n) {
    celltype <- categories[[n]][2]
    if (!is.null(celltype) && celltype %in% c("immune", "stromal", "epithelial")) {
        switch(celltype,
               "immune" = "circle",
               "stromal" = "square",
               "epithelial" = "sphere")
    } else {
        "circle"
    }
})

# Plot
set.seed(1)
layout <- layout_with_fr(network)
pdf(paste0(wd, "niches/patient4_only/network_", sample, "_manualColor.pdf"), width = 4, height = 4)
par(mar = c(2, 2, 2, 2))
plot(network,
    layout = layout,
    vertex.size = pmax(degree(network, mode="all") * 2.7, 4),
    vertex.color = cluster_colors,
    vertex.shape = node_shapes,
    vertex.frame.color = "white",
    vertex.label.cex = 0.5,
    vertex.label.family = "Helvetica",
    vertex.label.color = "black",
    edge.width = pmin(E(network)$weight * 3, 5),
    edge.color = rgb(0.6, 0.6, 0.6, 0.3),
    main = paste(sample, "- Spatial Niches (Manual Colors)")
)
dev.off()

# Also plot the spatial distribution of the maximum niche labels using the same manual colors
vis <- spls[[sample]]
niche_matrix <- GetAssayData(vis, assay = "niche_score", slot = "data")
max_features <- apply(niche_matrix, 2, function(x) {
    if (all(is.na(x))) {
        return(NA)
    } else {
        return(rownames(niche_matrix)[which.max(x)])
    }
})
# Fix dash/underscore issue for top_niche and manual_colors_patient4
vis$top_niche <- gsub("_", "-", unlist(max_features))
manual_colors_patient4_dash <- manual_colors_patient4
names(manual_colors_patient4_dash) <- gsub("_", "-", names(manual_colors_patient4))

if (length(vis$sample_id) > 5000) {
    pt_size <- 1.4
} else {
    pt_size <- 2.3
}
p <- MapLabels(
        vis,
        column_name = "top_niche",
        label_by = "sample_id",
        colors = manual_colors_patient4_dash,
        pt_size = pt_size,
        override_plot_dims = TRUE,
        drop_na = TRUE
    ) &
    theme(legend.position = "right") &
    guides(fill = guide_legend(override.aes = list(size = 4)))
ggsave(
    filename = paste0(wd, "niches/patient4_only/", sample, "_niche_labels_manualColor.jpg"),
    plot = p,
    height = 5,
    width = 7,
    dpi = 600,
    limitsize = FALSE
)
```

################################################################## Subpopulation miniatlas comparison ######################################################################

# load 
```{r}
library(igraph)
wd <- paste0(FIGS_ROOT, "/breast_cancer/analysis_cellStateLevel/ref_atlas_comparisons/")
int.all <- readRDS(paste0(rootObj, "integrated_all_subpopulations.rds"))
mergedVis <- readRDS(paste0(rootObj, "merged_visium_deconSubpopulations.rds"))
bc_atlas <- readRDS(paste0(DATA_ROOT, "/external_references/BC_atlas/miniatlas.rds"))
# split by sample
spls <- list()
samples <- unique(mergedVis$sample_id)
for (sample in samples){
    spl <- SubsetSTData(mergedVis, expression = sample_id == !!sample)
    spls[[sample]] <- spl
}
```

###### map Vis 55um with ref. atlas
```{r}
samples <- unique(int.all$sample[!grepl("HD", int.all$sample)])
bc_atlas$celltype_subset <- gsub("[ _]", "-", bc_atlas$celltype_subset)
features = unique(bc_atlas$celltype_subset)
for (sample in samples){
    vis <- spls[[sample]]
    patient_ID <- sub("_.*$", "", sample)
    ref <- bc_atlas
    
    ## Map w/ semla NNLS
    ident <- "celltype_subset"
    celltype <- "garvan_subpopulation"
    DefaultAssay(vis) <- "Spatial"
    DefaultAssay(ref) <- "RNA"
    vis <- vis |>
    NormalizeData() |>
    FindVariableFeatures(nfeatures = 2000) |>
    ScaleData()
    vis <- RunNNLS(vis, singlecell_object = ref, spatial_assay = DefaultAssay(vis), singlecell_assay = DefaultAssay(ref), groups = ident, slot = "data", assay_name = celltype)
    
    DefaultAssay(vis) <- celltype
    spls[[sample]] <- vis
}
```

###### plot each celltype and sample independently
```{r}
autoscale <- FALSE
img_filename <- "_refAtlas_vis55_2000varfeatures.jpg"
for (sample in samples) {
    vis <- spls[[sample]]
    DefaultAssay(vis) <- "garvan_subpopulation"
    subfolder <- if (autoscale) "autoscaled" else "scale_0_to_1"
    if (!dir.exists(paste0(wd, "map/individual_maps/", sample, "/", subfolder))) {
        dir.create(paste0(wd, "map/individual_maps/", sample, "/", subfolder), recursive = TRUE)
    }
        if (length(vis$sample_id) > 5000) {
        pt_size <- 0.7
    } else if (length(vis$sample_id) < 5000) {
        pt_size <- 1.2
    }
    for (feature in features){
            p <- MapFeatures(vis, features = feature, colors = viridis::turbo(11), label_by = "sample_id", pt_size = pt_size, override_plot_dims = TRUE, scale = "shared", ncol = 1)
            if (!autoscale) {
                p <- p & scale_fill_viridis_c(limits = c(0,1), option = "H") # make scale 0 to 1
            }
            ggsave(
                filename = paste0(wd, "map/", "individual_maps/", sample, "/", subfolder, "/", gsub("/", "_", feature), img_filename),
                plot = p,
                height = 4,
                width = 4,
                dpi = 600,
                limitsize = FALSE
            )
    }
}
```

##### merge Visium object
```{r}
spls[[3]]@assays <- spls[[3]]@assays[!grepl("^c2l_", names(spls[[3]]@assays))] # these assays are screwing up the merging 
spls[[4]]@assays <- spls[[4]]@assays[!grepl("^c2l_", names(spls[[4]]@assays))]

mergedVis <- MergeSTData(spls[[1]], spls[-1])
table(mergedVis$sample_id)
# saveRDS(mergedVis, paste0(rootObj, "merged_visium_deconSubpopulations_simplex.rds"))
```

##### merged
```{r}
DefaultAssay(mergedVis) <- "garvan_subpopulation"
# Fetch data and compute correlation
data <- FetchData(mergedVis, rownames(mergedVis)) %>% 
    as.data.frame() %>%
    select(where(~ sum(., na.rm = TRUE) > 0))  # Add na.rm=TRUE to handle NAs
cor_matrix <- cor(data, method = "spearman", use = "pairwise.complete.obs")
diag(cor_matrix) <- NA

# Generate heatmap and save as PDF
pdf(paste0(wd, "correlation/correlation_bc_atlas_subset.pdf"), width = 13, height = 10)
heatmap3(
    cor_matrix, 
    margins = c(14, 14), 
    symm = TRUE,
    col = colorRampPalette(c("blue", "white", "red"))(20)
)

dev.off()
# Save correlation matrix as CSV
write.csv(cor_matrix, file = paste0(wd, "correlation/correlation_matrix.csv"), row.names = TRUE)
```

```{r}
# Create a vector with matches between sample and subtype
patient_subtypes <- c(
    "patient1_55um" = "LuminalA",
    "patient2_55um" = "LuminalB(ER)",
    "patient3_55um" = "TNBC",
    "patient4_55um" = "TNBC",
    "patient4_HD" = "TNBC",
    "patient5_55um" = "TNBC",
    "patient5_HD" = "TNBC",
    "patient6_55um" = "TNBC",
    "patient7_55um" = "LuminalB(ER)",
    "patient8_55um" = "LuminalA",
    "patient9_55um" = "LuminalB(ER)"
)
sbtps <- list()
subtypes <- unique(patient_subtypes)
for (subtype in subtypes){
    sbt <- SubsetSTData(mergedVis, expression = sample_id %in% names(patient_subtypes[patient_subtypes == subtype]))
    sbtps[[subtype]] <- sbt
}
```

###### defining colors for each type
```{r}
# Define categories based on patient or subtype information
categories <- list(
    "B-cells-Memory" = c("NA", "immune"),
    "B-cells-Naive" = c("NA", "immune"),
    "CAFs-MSC-iCAF-like-s1" = c("NA", "stromal"),
    "CAFs-MSC-iCAF-like-s2" = c("NA", "stromal"),
    "CAFs-Transitioning-s3" = c("NA", "stromal"),
    "CAFs-myCAF-like-s4" = c("NA", "stromal"),
    "CAFs-myCAF-like-s5" = c("NA", "stromal"),
    "Cancer-Basal-SC" = c("NA", "epithelial"),
    "Cancer-Cycling" = c("NA", "epithelial"),
    "Cancer-Her2-SC" = c("NA", "epithelial"),
    "Cancer-LumA-SC" = c("NA", "epithelial"),
    "Cancer-LumB-SC" = c("NA", "epithelial"),
    "Cycling-Myeloid" = c("NA", "immune"),
    "Cycling-PVL" = c("NA", "stromal"),
    "Endothelial-ACKR1" = c("NA", "stromal"),
    "Endothelial-CXCL12" = c("NA", "stromal"),
    "Endothelial-Lymphatic-LYVE1" = c("NA", "stromal"),
    "Endothelial-RGS5" = c("NA", "stromal"),
    "Luminal-Progenitors" = c("NA", "epithelial"),
    "Mature-Luminal" = c("NA", "epithelial"),
    "Myeloid-c0-DC-LAMP3" = c("NA", "immune"),
    "Myeloid-c1-LAM1-FABP5" = c("NA", "immune"),
    "Myeloid-c10-Macrophage-1-EGR1" = c("NA", "immune"),
    "Myeloid-c11-cDC2-CD1C" = c("NA", "immune"),
    "Myeloid-c12-Monocyte-1-IL1B" = c("NA", "immune"),
    "Myeloid-c2-LAM2-APOE" = c("NA", "immune"),
    "Myeloid-c3-cDC1-CLEC9A" = c("NA", "immune"),
    "Myeloid-c4-DCs-pDC-IRF7" = c("NA", "immune"),
    "Myeloid-c5-Macrophage-3-SIGLEC1" = c("NA", "immune"),
    "Myeloid-c7-Monocyte-3-FCGR3A" = c("NA", "immune"),
    "Myeloid-c8-Monocyte-2-S100A9" = c("NA", "immune"),
    "Myeloid-c9-Macrophage-2-CXCL10" = c("NA", "immune"),
    "Myoepithelial" = c("NA", "epithelial"),
    "PVL-Differentiated-s3" = c("NA", "stromal"),
    "PVL-Immature-s1" = c("NA", "stromal"),
    "PVL-Immature-s2" = c("NA", "stromal"),
    "Plasmablasts" = c("NA", "immune"),
    "T-cells-c0-CD4+-CCR7" = c("NA", "immune"),
    "T-cells-c1-CD4+-IL7R" = c("NA", "immune"),
    "T-cells-c10-NKT-cells-FCGR3A" = c("NA", "immune"),
    "T-cells-c11-MKI67" = c("NA", "immune"),
    "T-cells-c2-CD4+-T-regs-FOXP3" = c("NA", "immune"),
    "T-cells-c3-CD4+-Tfh-CXCL13" = c("NA", "immune"),
    "T-cells-c4-CD8+-ZFP36" = c("NA", "immune"),
    "T-cells-c5-CD8+-GZMK" = c("NA", "immune"),
    "T-cells-c6-IFIT1" = c("NA", "immune"),
    "T-cells-c7-CD8+-IFNG" = c("NA", "immune"),
    "T-cells-c8-CD8+-LAG3" = c("NA", "immune"),
    "T-cells-c9-NK-cells-AREG" = c("NA", "immune")
)

# Define corresponding colors for each category
category_colors <- list(
    specificity = c(
"NA"
    ),
    celltype = c(
        "immune" = "#3498DB",
        "stromal" = "#2ECC71",
        "epithelial" = "#E74C3C"
    )
)
```

#### per subtype
```{r}
recolored <- TRUE # we go back to these plots after recoloring in downstream analysis 
subtypeNetworks <- list()
for (subtype in subtypes){
    sbt <- sbtps[[subtype]]
    DefaultAssay(sbt) <- "garvan_subpopulation"
    data <- FetchData(sbt, rownames(sbt)) %>% 
        as.data.frame() %>%
        select(where(~ sum(., na.rm = TRUE) > 0))  # Add na.rm=TRUE to handle NAs

    cor_matrix <- cor(data, method = "spearman", use = "pairwise.complete.obs")
    diag(cor_matrix) <- NA

    set.seed(1)
    corMat_network <- cor_matrix |> t()
    corMat_network[corMat_network<0.2] <- 0
    network <- graph_from_adjacency_matrix(corMat_network, weighted=T, mode="undirected", diag=F)

    # Count the number of degree for each node:
    deg <- degree(network, mode="all")

    # Assign colors based on celltype
    node_names <- V(network)$name
    celltype_colors <- sapply(node_names, function(n) {
        ct <- categories[[n]][2]  # 2nd element is "celltype"
        if (!is.null(ct) && ct %in% names(category_colors$celltype)) {
            return(category_colors$celltype[[ct]])
        } else {
            return("#BDC3C7")  # fallback color (light grey)
        }
    })

    subtypeNetworks[[subtype]] <- network
    
    # Plot
    layout <- layout_with_fr(network)
    if (recolored) {
        pdf(paste0(wd, "niche_analysis/networkFinal_subtype_", subtype, "_colorBy_celltype.pdf"), width = 7, height = 7)
    } else {
        pdf(paste0(wd, "correlation/network_visualization_subtype_", subtype, ".pdf"), width = 7, height = 7)
    }
    par(mar = c(0, 0, 1, 1))  # Broaden right margin
    plot(network,
        layout = layout,
        vertex.size = pmax(deg * 2.7, 4),  # minimum size for visibility
        vertex.color = celltype_colors,
        vertex.frame.color = "white",
        vertex.label.cex = 0.7,
        vertex.label.family = "Helvetica",
        vertex.label.color = "black",
        edge.width = pmin(E(network)$weight * 3, 5),  # cap edge width
        edge.color = rgb(0.6, 0.6, 0.6, 0.3),  # light grey transparent edges
        main = paste(subtype))
    # Add legend for celltype colors
    legend(
        "topright",  
        legend = names(category_colors$celltype),  
        fill = category_colors$celltype,  
        title = "CellType",
        inset = c(0, 0)  # Adjust inset further to ensure it fits within the plot area
    )
    dev.off()
}
```

### Cluster based on graph connectivity - louvain algorithm
```{r}
# Function to get clusters for each subtype network
get_clusters <- function(network) {
    clusters <- cluster_louvain(network)
    membership <- membership(clusters)
    return(membership)
}
# Apply clustering to each subtype network
subtype_clusters <- lapply(subtypeNetworks, get_clusters)
if (recolored == FALSE) {
    niche_colors <- setNames(rainbow(length(unique(unlist(subtype_clusters)))), unique(unlist(subtype_clusters)))
}
# Add subtype clusters to graph plot
for (subtype in names(subtype_clusters)) {
    network <- subtypeNetworks[[subtype]]
    membership <- subtype_clusters[[subtype]]
    
    # Assign colors based on niche membership and subtype
    cluster_colors <- sapply(names(membership), function(n) {
        if (recolored) {
            subtype_name <- subtype  # Use the current subtype being iterated
            niche_name <- paste0(subtype_name, "-niche", membership[n])
            if (niche_name %in% names(niche_colors) && sum(membership == membership[n]) > 1) {
                return(niche_colors[niche_name])
            } else {
                return("#BDC3C7")  # fallback color (light grey)
            }
        } else {
            if (sum(membership == membership[n]) > 1) {
                rainbow(length(unique(membership)))[membership[n]]
            } else {
                "#BDC3C7"  # fallback color for single-member clusters
            }
        }
    })
    
    # Assign shapes based on cell type
    node_shapes <- sapply(names(membership), function(n) {
        celltype <- categories[[n]][2]  # Get cell type from categories
        if (!is.null(celltype) && celltype %in% c("immune", "stromal", "epithelial")) {
            switch(celltype,
                   "immune" = "circle",
                   "stromal" = "square",
                   "epithelial" = "sphere")
        } else {
            "circle"  # Default shape
        }
    })
    
    # Plot
    set.seed(1)  # Ensure reproducibility of layout
    layout <- layout_with_fr(network)
    if (recolored) {
        pdf(paste0(wd, "niche_analysis/networkFinal_subtype_", subtype, "_colorBy_niche.pdf"), width = 9, height = 7)
    } else {
        pdf(paste0(wd, "correlation/network_visualization_subtype_", subtype, "_clusters.pdf"), width = 9, height = 7)
    }
    par(mar = c(0, 0, 1, 2))  # Broaden right margin
    plot(network,
        layout = layout,
        vertex.size = pmax(degree(network, mode="all") * 2.7, 4),  # minimum size for visibility
        vertex.color = cluster_colors,
        vertex.shape = node_shapes,
        vertex.frame.color = "white",
        vertex.label.cex = 0.7,
        vertex.label.family = "Helvetica",
        vertex.label.color = "black",
        edge.width = pmin(E(network)$weight * 3, 5),  # cap edge width
        edge.color = rgb(0.6, 0.6, 0.6, 0.3),  # light grey transparent edges
        main = paste(subtype, "- Spatial Niches"))
    
    # Add legend for clusters
    legend(
        "topright",  
        legend = c(paste("Niche", unique(membership[cluster_colors != "#BDC3C7"])), "No niche"),  
        fill = c(sapply(unique(membership[cluster_colors != "#BDC3C7"]), function(n) niche_colors[paste0(subtype, "-niche", n)]), "#BDC3C7"),  
        title = "Niches",
        inset = c(0, 0), 
        border = NA
    )
        # Add legend for shapes
    legend(
        "bottomright",
        legend = c("Immune", "Stromal", "Epithelial"),
        pch = c(21, 22, 16),  # Corresponding shapes
        title = "Cell Lineage",
        inset = c(0, 0),
        col = "black",
        pt.bg = "white"
    )
    dev.off()
}
```

## Create niche scores based on graph clusters
```{r}
niches <- list()
for (subtype in names(subtype_clusters)) {
    clusters <- subtype_clusters[[subtype]]
    unique_clusters <- unique(clusters)
    for (cluster_id in unique_clusters) {
        cluster_members <- names(clusters[clusters == cluster_id])
        celltypes <- unique(sapply(cluster_members, function(n) categories[[n]][2]))
        if (length(celltypes) > 1) {
            niche_name <- paste0(subtype, "_niche", cluster_id)
            niches[[niche_name]] <- cluster_members
        }
    }
}
subpop_df <- as.data.frame(t(GetAssayData(mergedVis, assay = "garvan_subpopulation", slot = "data")))
niche_scores_list <- list()

# For each niche
for (niche_name in names(niches)) {
    cell_types <- niches[[niche_name]]
    
    # Filter only the columns in subpop_df that match cell types in this niche
    common_cell_types <- intersect(cell_types, colnames(subpop_df))
    
    if (length(common_cell_types) > 0) {
        # Extract relevant columns
        niche_matrix <- subpop_df[, common_cell_types, drop = FALSE]
        
        # Compute row-wise sum of log-transformed proportions
        niche_score <- rowSums(niche_matrix)

        # Store the vector
        niche_scores_list[[niche_name]] <- niche_score
    } else {
        warning(paste("No matching cell types found for", niche_name))
        # Fill with NA if no cell types matched
        niche_scores_list[[niche_name]] <- rep(NA, nrow(subpop_df))
    }
}

# Combine all niche scores into a data frame
niche_score_mat <- as.data.frame(niche_scores_list, check.names = FALSE)
rownames(niche_score_mat) <- rownames(subpop_df)

# Transpose to get niches as rows (optional depending on downstream analysis)
niche_score_mat <- t(as.matrix(niche_score_mat))

# Z-score normalization across samples for each niche (row-wise)
niche_score_mat <- t(scale(t(niche_score_mat)))  # scales across columns (samples)

mergedVis[["niche_score_garvan"]] <- CreateAssayObject(data = niche_score_mat)
```

### map niche scores spatially
###### plot each niche score and sample independently
```{r}
autoscale <- TRUE
img_filename <- "_nicheScore.jpg"
DefaultAssay(mergedVis) <- "niche_score_garvan"
features <- rownames(mergedVis)
for (sample in samples) {
    vis <- SubsetSTData(mergedVis, expression = sample_id == !!sample)
    subfolder <- if (autoscale) "autoscaled" else "scale_0_to_1"
    if (!dir.exists(paste0(wd, "niche_map/individual_maps/", sample, "/", subfolder))) {
        dir.create(paste0(wd, "niche_map/individual_maps/", sample, "/", subfolder), recursive = TRUE)
    }
        if (length(vis$sample_id) > 5000) {
        pt_size <- 0.7
    } else if (length(vis$sample_id) < 5000) {
        pt_size <- 1.2
    }
    for (feature in features){
            p <- MapFeatures(vis, features = feature, colors = viridis::turbo(11), label_by = "sample_id", pt_size = pt_size, override_plot_dims = TRUE, scale = "shared", ncol = 1)
            if (!autoscale) {
                p <- p & scale_fill_viridis_c(limits = c(0,1), option = "H") # make scale 0 to 1
            }
            if (!dir.exists(paste0(wd, "niche_analysis/", "map_individual/", sample, "/", subfolder))) {
                dir.create(paste0(wd, "niche_analysis/", "map_individual/", sample, "/", subfolder), recursive = TRUE)
            }
            ggsave(
                filename = paste0(wd, "niche_analysis/", "map_individual/", sample, "/", subfolder, "/", gsub("/", "_", feature), img_filename),
                plot = p,
                height = 4,
                width = 4,
                dpi = 600,
                limitsize = FALSE
            )
    }
}
```

## Create discrete Niche labels in spatial data

###### define colors for each niche
```{r}
niche_colors <- list(
    # Luminal A - bold cool colors
    "LuminalA-niche1" = paletteer_d("ggthemes::excel_Blue")[2],
    "LuminalA-niche3" = paletteer_d("ggthemes::excel_Blue")[3],  
    "LuminalA-niche4" = paletteer_d("ggthemes::excel_Blue")[4],  
    "LuminalA-niche12" = paletteer_d("ggthemes::excel_Blue")[5],

    # Luminal B(ER) - bold purples/pinks
    "LuminalB(ER)-niche1" = paletteer_d("ggthemes::excel_Violet_II")[1],  
    "LuminalB(ER)-niche3" = paletteer_d("ggthemes::excel_Violet_II")[3], 
    "LuminalB(ER)-niche4" = paletteer_d("ggthemes::excel_Violet_II")[4], 
    "LuminalB(ER)-niche5" = paletteer_d("ggthemes::excel_Violet_II")[5], 
    "LuminalB(ER)-niche9" = paletteer_d("ggthemes::excel_Violet_II")[6],

    # TNBC - bold fiery spectrum
    "TNBC-niche1" = paletteer_d("RColorBrewer::YlOrRd")[8], 
    "TNBC-niche3" = paletteer_d("RColorBrewer::YlOrRd")[6], 
    "TNBC-niche6" = paletteer_d("RColorBrewer::YlOrRd")[7], 
    "TNBC-niche7" = paletteer_d("RColorBrewer::YlOrRd")[4], 
    "TNBC-niche8" = paletteer_d("RColorBrewer::YlOrRd")[5]   
)
niche_colors <- unlist(niche_colors)
```

###### Create maximum value metadata labels for the niche scores
```{r}
niche_matrix <- GetAssayData(mergedVis, assay = "niche_score_garvan", slot = "data")

# Get the feature (row) with the max value per column (spot)
max_features <- apply(niche_matrix, 2, function(x) {
  if (all(is.na(x))) {
    return(NA)  # No niche passes threshold
  } else {
    return(rownames(niche_matrix)[which.max(x)])
  }
})
mergedVis$top_niche_garvan <- unlist(max_features)
```

##### plot labels
```{r}
samples <- unique(mergedVis$sample_id)
for (sample in samples) {
    vis <- SubsetSTData(mergedVis, expression = sample_id == !!sample)
    vis <- LoadImages(vis)
    if (length(vis$sample_id) > 5000) {
        pt_size <- 1.4
    } else {
        pt_size <- 2.3
    }
    p <- MapLabels(vis, column_name = "top_niche_garvan", label_by = "sample_id", colors = niche_colors, pt_size = pt_size, override_plot_dims = TRUE, image_use = "raw", drop_na = TRUE) &
        theme(legend.position = "right") &
        guides(fill = guide_legend(override.aes = list(size = 4)))
    ggsave(
        filename = paste0(wd, "niche_analysis/map_labels/", sample, "_niche_labels.jpg"),
        plot = p,
        height = 5,
        width = 7,
        dpi = 600,
        limitsize = FALSE
    )
}
```

### apply niche coloring scheme to subtype-specific network plots
we go back to previous code and set "recolored" to TRUE


################################################################## HD mapping #####################################################################################################

### plot diff references to HD data for patient 5 

```{r}
wd <- paste0(FIGS_ROOT, "/breast_cancer/analysis_cellStateLevel/HD/")
```

```{r}
patient <- "patient5"
ress <- c("s_002um", "s_008um", "s_016um", "s_048um")
simp <- subset(int.all, subset = patient_ID == patient)
bc_atlas$subpopulation <- bc_atlas$celltype_subset
refs <- list(bc_atlas = bc_atlas, SIMPlex = simp)

vis <- readRDS(paste0(DATA_ROOT, "/r_objects/breast_cancer/visium/HD/patient5_HD/patient5_HD_spatial.rds"))
subset48 <- SubsetSTData(vis, spots = grep("s_048um", colnames(vis), value = TRUE)) 
    DefaultAssay(subset48) <- "Spatial"
    subset48 <- subset48 |>
    NormalizeData() |>
    FindVariableFeatures(nfeatures = 2000) |>
    ScaleData()
var.features <- VariableFeatures(subset48)

resObj <- list()
for (res in ress) {
    barcodes_in_range <- grep(paste0("^", res), colnames(vis), value = TRUE)
    if (res == "s_002um") {
        barcodes_in_range <- vis@tools$Staffli@meta_data %>%
            filter(pxl_col_in_fullres >= max(pxl_col_in_fullres * 0.4), # left
                pxl_col_in_fullres <= max(pxl_col_in_fullres * 0.6), # right 
                pxl_row_in_fullres >= max(pxl_row_in_fullres * 0.45), # top
                pxl_row_in_fullres <= max(pxl_row_in_fullres * 0.65)) %>% # bottom
            pull(barcode) %>%
            grep(paste0("^", res), ., value = TRUE)
    }
    subRes <- SubsetSTData(vis, spots = barcodes_in_range)
    ## quick check of the frame
    ggsave(paste0(wd, "map_subtypes/patient5_area.jpg"), MapFeatures(subRes, features = "nFeature_Spatial", override_plot_dims = TRUE, pt_size = 0.5, shape = "tile"))
    ## Map w/ semla NNLS
    ident <- "subpopulation" # ident in the reference to use for mapping 
    DefaultAssay(subRes) <- "Spatial"
        subRes <- subRes |>
        NormalizeData() |>
        ScaleData()
    VariableFeatures(subRes) <- var.features # we use bin 48's var features due to sparsity of HD 

    for (refName in c("bc_atlas", "SIMPlex")){
        celltype <- paste0(refName, "_subpopulation_map_", res) # name of the mapping to assign to the visium object
        ref <- refs[[refName]]
        ref$subpopulation <- gsub("/", "-", ref$subpopulation)
        DefaultAssay(subRes) <- "Spatial"
        subRes <- RunNNLS(subRes, singlecell_object = ref, spatial_assay = DefaultAssay(subRes), singlecell_assay = DefaultAssay(ref), groups = ident, slot = "data", assay_name = celltype)
        DefaultAssay(subRes) <- celltype
        features <- unique(gsub("_", "-", rownames(subRes)))
        features <- features[order(features)]  # Ensure consistent order

        # plot image per resolution per reference per celltype
        for (feature in features) {
            p <- MapFeatures(subRes, features = feature, colors = viridis::turbo(11), label_by = "sample_id", override_plot_dims = TRUE, scale = "shared", ncol = 1, shape = "tile", drop_na = TRUE)
            path <- paste0(wd, "map_subtypes/individual_plots/", res, "/", refName, "/")
            if (!dir.exists(path)) {
                dir.create(path, recursive = TRUE)
            }
            ggsave(paste0(path, patient, "_", feature, "_mapHD.jpg"), p, height = 4, width = 4, dpi = 300, limitsize = FALSE)
        }
    }
    resObj[[res]] <- subRes
}
```

### plot maximum celltype assignment per datapoint @16um
```{r}
res <- "s_016um"
crop <- FALSE
subRes <- resObj[[res]]
barcodes_in_range <- grep(paste0("^", res), colnames(subRes), value = TRUE)
if (crop == TRUE) {
    barcodes_in_range <- subRes@tools$Staffli@meta_data %>%
        filter(pxl_col_in_fullres >= max(pxl_col_in_fullres * 0.4), # left
            pxl_col_in_fullres <= max(pxl_col_in_fullres * 0.6), # right 
            pxl_row_in_fullres >= max(pxl_row_in_fullres * 0.45), # top
            pxl_row_in_fullres <= max(pxl_row_in_fullres * 0.65)) %>% # bottom
        pull(barcode) %>%
        grep(paste0("^", res), ., value = TRUE)
}
subRes <- SubsetSTData(subRes, spots = barcodes_in_range)

for (refName in c("bc_atlas", "SIMPlex")){
    # Plot each resolution separately
    features <- unique(gsub("_", "-", rownames(subRes)))
    features <- features[order(features)]  # Ensure consistent order
    DefaultAssay(subRes) <- paste0(refName, "_subpopulation_map_s_016um")

        # get max value 
        mtx <- GetAssayData(subRes, assay = paste0(refName, "_subpopulation_map_s_016um"), slot = "data")
        max_features <- apply(mtx, 2, function(x) {
        if (all(is.na(x))) {
            return(NA)  # No niche passes threshold
        } else {
            return(rownames(mtx)[which.max(x)])
        }
        })
        subRes$top_celltype <- unlist(max_features)

    # cols <- paletteer_c("ggthemes::Sunset-Sunrise Diverging", n = length(unique(subRes$top_celltype)))
    p <- MapLabels(subRes, column_name = "top_celltype", override_plot_dims = TRUE, shape = "tile", drop_na = TRUE) &
  theme(legend.position = "right")
    if (crop) {
        ggsave(paste0(wd, "map_subtypes/top_subtype/", refName,"_", patient , "_discretized_mapHD_zoomed.jpg"), p, height = 10, width = 15, dpi = 300, limitsize = FALSE)
    } else {
        ggsave(paste0(wd, "map_subtypes/top_subtype/", refName,"_", patient , "_discretized_mapHD.jpg"), p, height = 10, width = 15, dpi = 300, limitsize = FALSE)
    }
}
```

```{r}
# saveRDS(resObj, paste0(rootObj, "HD_patient5_decon_atlas_simplex_subpopulations.rds"))
```

################################################################## Specific analyses #####################################################################################################

# assess composition radial composition of DCIS vs INVASIVE 
in order to assess the cell state composition of DCIS and INVASIVE areas we 
do a radial analysis with the segmentation of the DCIS and invasive histopathology 
labels in patient 4 


## load 
```{r}
int.all <- readRDS(paste0(DATA_ROOT, "/r_objects/breast_cancer/single_nuclei/cross_patient/integrated_all_subpopulations.rds"))
mergedVis <- readRDS(paste0(rootObj, "merged_visium_deconSubpopulations_simplex.rds"))
rootObj <- paste0(DATA_ROOT, "/r_objects/breast_cancer/single_nuclei/cross_patient/")
wd <- paste0(FIGS_ROOT, "/breast_cancer/analysis_cellStateLevel/pat4_invasive_dcis_composition/")

## pat4 only obj
sample <- "patient4_55um"
pat4 <- SubsetSTData(mergedVis, sample_id %in% sample)
```

## population diversity surounding invasive niches 

### load path annotation for TNBC samples
```{r}
types <- c("epithelial", "immune", "stroma", "additional")
annot <- list()
for (type in types) {
    csv_path <- file.path(here::here("resources/breast_cancer/visium_histpathology"), sample, paste0(type, ".csv"))
    annot[[type]] <- read.csv(csv_path)

    rownames(annot[[type]]) <- annot[[type]]$Barcode
    annot[[type]] <- annot[[type]][rownames(annot[[type]]) %in% rownames(pat4@meta.data), ]
    annot[[type]][annot[[type]] == ""] <- NA  # Replace empty cells with NA
    pat4 <- AddMetaData(pat4, metadata = annot[[type]], col.name = colnames(annot[[type]])[2])
}
p_list <- list()
for (type in types) {
    p_list[[type]] <- MapLabels(pat4, column_name = type, ncol = 1, label_by = "sample_id", override_plot_dims = TRUE, drop_na = TRUE, pt_size = 0.9)
}
p <- patchwork::wrap_plots(p_list, ncol = length(types)) & guides(fill = guide_legend(override.aes = list(size = 6), ncol = 1)) & theme(legend.position = "right")
ggsave(paste0(wd, sample, "_annotation.jpg"), p, height = 5, width = length(types) * 5, dpi = 600, limitsize = FALSE)
```

#### merge invasive dense and invasive sparse
```{r}
pat4$epithelial[pat4$epithelial %in% c("invasive_dense", "invasive_sparse")] <- "invasive"
pat4$epithelial[pat4$epithelial == "in_situ_dcis"] <- "dcis"
```

## radial plot for cancer areas in patient 4 

#### get radial distances for top niche
```{r}
groups <- c("invasive", "dcis")
for (group in groups){
pat4 <- RadialDistance(pat4, column_name = "epithelial", selected_groups = groups, convert_to_microns = TRUE)
pat4$rooted_dist_dcis <- sign(pat4$r_dist_dcis)*sqrt(abs(pat4$r_dist_dcis))
pat4$rooted_dist_invasive <- sign(pat4$r_dist_invasive)*sqrt(abs(pat4$r_dist_invasive))
p <- MapFeatures(pat4, features = c(paste0("rooted_dist_", groups)), pt_size = 1, ncol = 2, drop_na = TRUE, label_by = "sample_id",
            colors = RColorBrewer::brewer.pal(n = 11, name = "RdBu") |> rev(),
            override_plot_dims = TRUE)
ggsave(
    filename = paste0(wd, paste0("rooted_distances.jpg")),
    plot = p,
    height = 5,
    width = 10,
    dpi = 600,
    limitsize = FALSE
)
}
```

#### radial distance stacked plot 

##### color map
```{r}
color_map <- c(
    # Adipocytes
    "Adipocytes" = "#A9A9A9",  # gray

    # Cancer populations (vibrant reds)
    "Cancer-Basal/Plastic-pat4"   = "#FF9900", # orange-red (distinct from deep red)
    "Cancer-Transitional-pat4"    = "#FF6347", # tomato (distinct, more orange)
    "Cancer-Basal/Cycl-pat4"      = "#B20000", # deep red

    # Luminal progenitors (oranges)
    "LumProgenitors-Cycl-pat4"       = "#27AE60", # green
    "LumProgenitors-transitional-pat4" = "#82E0AA", # light green

    # CAF populations
    # apCAF (purples)
    "apCAF-classical"       = "#8E44AD", # purple
    # iCAF (yellow)
    "iCAF-classical"        = "#FFE066", # yellow
    "iCAF-lipid-associated" = "#e4cf7bff", # light yellow
    "iCAF-secretory"        = "#c4a310ff", # pale yellow
    # mCAF (deep reds)
    "mCAF-classical"        = "#800000", # deep maroon
    "mCAF-patient4-specific" = "#A93226", # deep red

    # Immune (blue tones)
    "Macrophages-immunoregulatory" = "#6BAED6", # sky blue
    "Macrophages-inflammatory"     = "#2171B5", # strong medium blue
    "B/T-mixed-pat4"               = "#9ECAE1", # pale blue
    "pDCs"                         = "#08519C",  # dark navy blue
    "B/T-mixed-patient4"           = "#C6DBEF" # very pale blue
)
```
```{r}
DefaultAssay(pat4) <- "subpopulation"
sel_genes <- rownames(pat4)
# subpop_colors <- setNames(rainbow(length(sel_genes)), sel_genes)

for (radial_distance_col in c("r_dist_dcis", "r_dist_invasive")) {
    pdat <- FetchData(pat4, vars = c(sel_genes, radial_distance_col))

    # Bin the x-axis (radial distance) for smoothing
    bin_width <- 100  # adjust as needed for smoothing
    pdat$bin_r_dist <- round(pdat[[radial_distance_col]] / bin_width) * bin_width

    # Prepare long format and filter range
    pdat_long <- pdat %>%
        filter(.data[[radial_distance_col]] > -1000 & .data[[radial_distance_col]] < 2000) %>%
        pivot_longer(all_of(sel_genes), names_to = "variable", values_to = "value")

    # Remove Plasma_cells or Plasma-cells from the variable column
    pdat_long <- pdat_long %>%
        filter(!grepl("^Plasma[_-]?cells$", variable, ignore.case = TRUE))

    # Calculate mean for positive and negative distances for each subpopulation
    diff_df <- pdat_long %>%
        mutate(region = ifelse(.data[[radial_distance_col]] >= 0, "positive", "negative")) %>%
        group_by(variable, region) %>%
        summarise(mean_value = mean(value, na.rm = TRUE), .groups = "drop") %>%
        pivot_wider(names_from = region, values_from = mean_value, values_fill = 0) %>%
        mutate(diff = abs(positive - negative))

    # Select top 10 subpopulations with largest difference
    top_vars <- diff_df %>%
        arrange(desc(diff)) %>%
        slice_head(n = 10) %>%
        pull(variable)

    # Filter long data to only top subpopulations
    pdat_long_top <- pdat_long %>%
        filter(variable %in% top_vars) %>%
        group_by(bin_r_dist, variable) %>%
        summarise(value = mean(value, na.rm = TRUE), .groups = "drop")

    p <- ggplot(pdat_long_top, aes(x = bin_r_dist, y = value, fill = variable)) +
        geom_area(position = "fill") +
        geom_vline(xintercept = 0, linetype = "dashed", color = "#787171", size = 1) +
        scale_fill_manual(values = color_map, drop = TRUE) + 
        annotate("text", x = 20, y = 1.02, label = ifelse(radial_distance_col == "r_dist_invasive", "  Invasive border", "  DCIS border"), vjust = 0, hjust = 0, size = 4) +
        scale_x_continuous(name = "Distance (µm)") +
        scale_y_continuous(name = "Proportion") +
        theme_classic()

    ggsave(
        filename = paste0(wd, "radial_composition_", radial_distance_col, "_stacked.pdf"),
        plot = p,
        height = 5,
        width = 10,
        dpi = 600,
        limitsize = FALSE
    )
}
```

# Exploration of patient interesting population gene signatures
```{r}
wd <- paste0(FIGS_ROOT, "/breast_cancer/analysis_cellStateLevel/deg_interesting_celltypes/")   
```

```{r}
library(ggrepel)
# Volcano plot function with gene labels
volcano_compare <- function(seurat_obj, meta_col, group1, group2, 
                            logfc.threshold = 0.25, min.pct = 0.1, 
                            top_n = 10) {
  
  # Make sure meta_col exists
  if (!meta_col %in% colnames(seurat_obj@meta.data)) {
    stop(paste("Column", meta_col, "not found in metadata."))
  }
  
  # Remove "DEPRECATED-" genes from the RNA assay
  genes <- rownames(seurat_obj)
  keep_genes <- !grepl("^DEPRECATED-", genes)
  seurat_obj <- subset(seurat_obj, features = genes[keep_genes])
  
  # Set identities based on metadata column
  Idents(seurat_obj) <- seurat_obj@meta.data[[meta_col]]
  
  # Differential expression between two groups
  degs <- FindMarkers(
    seurat_obj,
    ident.1 = group1,
    ident.2 = group2,
    logfc.threshold = logfc.threshold,
    min.pct = min.pct,
    test.use = "wilcox"
  )
  
  # Add gene column
  degs$gene <- rownames(degs)
  
  # Select top genes for labeling:
  # - by adjusted p-value
  # - by absolute log2FC
  top_genes <- bind_rows(
    degs %>% arrange(p_val_adj) %>% head(top_n),
    degs %>% arrange(desc(abs(avg_log2FC))) %>% head(top_n)
  ) %>% distinct(gene, .keep_all = TRUE)
  
  # Volcano plot
  p <- ggplot(degs, aes(x = avg_log2FC, y = -log10(p_val_adj))) +
    geom_point(aes(color = avg_log2FC > 0), alpha = 0.7) +
    scale_color_manual(values = c("TRUE" = "firebrick", "FALSE" = "royalblue"),
                       labels = c(group2, group1)) +
    geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "gray50") +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "gray50") +
    geom_text_repel(
      data = top_genes,
      aes(label = gene),
      size = 3.5,
      max.overlaps = Inf,
      box.padding = 0.5,
      point.padding = 0.3,
      segment.color = "gray50"
    ) +
    theme_classic(base_size = 14) +
    labs(
      title = paste("Differential expression:", group1, "vs", group2),
      x = "Average log2 fold change",
      y = "-log10(adj. p-value)",
      color = "Enriched in"
    )
  
  return(list(DEGs = degs, plot = p))
}
```

```{r}
result_mcaf <- volcano_compare(int.all, meta_col = "subpopulation",
                          group2 = "mCAF_classical", group1 = "mCAF_patient4-specific")
ggsave(paste0(wd, "mCAF_patient4-specific_vs_mCAF_classical_volcano.pdf"), plot = result_mcaf$plot, width = 8, height = 4)
```

```{r}
result_lum <- volcano_compare(int.all, meta_col = "subpopulation",
                          group2 = "LumProgenitors_shared", group1 = "LumProgenitors_transitional_pat4")
ggsave(paste0(wd, "LumProgenitors_pat4_volcano.pdf"), plot = result_lum$plot, width = 8, height = 4)
```

```{r}
result_icaf <- volcano_compare(int.all, meta_col = "subpopulation",
                          group2 = "iCAF_classical", group1 = "iCAF_secretory")
ggsave(paste0(wd, "iCAF_secretory_volcano.pdf"), plot = result_icaf$plot, width = 8, height = 4)
```

Based on these marker gene profiles, we draw potential interactions in pat4 
worth to explore further in HD. the interaction of secretory iCAF with 
a potential transitional profile of luminal progenitors, and 
intra-invasive mCAFs that are patient specific. 
The border like profile of apCAFs is also worth investigating in HD.

# Spatially informed ligand receptor analysis iCAF secretory and luminal progenitors transitional
```{r}
library(celltalker)
int.all <- readRDS(paste0(DATA_ROOT, "/r_objects/breast_cancer/single_nuclei/cross_patient/integrated_all_subpopulations.rds"))
wd <- paste0(FIGS_ROOT, "/breast_cancer/analysis_cellStateLevel/ligand_receptor/")
```

#### patient 4 only 
```{r}
pat4_sn <- subset(int.all, subset = patient_ID == "patient4")
pat4_sn$subpopulation <- gsub("_", "-", pat4_sn$subpopulation) # celltalker has an issue with underscores
```
```{r}
## Run celltalker
interactions <- celltalk(input_object=pat4_sn,
  metadata_grouping="subpopulation",
  ligand_receptor_pairs=ramilowski_pairs,
  number_cells_required=30,
  min_expression=200,
  max_expression=20000,
  scramble_times=10)
# saveRDS(interactions, paste0(rootObj, "pat4_celltalker_interactions.rds"))
```

### spatially informed filtering
```{r}
mtx <- read.csv(paste0(FIGS_ROOT, "/breast_cancer/analysis_cellStateLevel/final_subpopulations/correlation/patient4_55um_correlation_matrix.csv", row.names = 1, check.names = FALSE)
mtx <- pmax(mtx, t(mtx)) # get bidirectional correlation
spatial_long <- as.data.frame(as.table(as.matrix(mtx))) %>%
  setNames(c("cell_type1", "cell_type2", "spatial_corr")) 

# Filter interactions
top_stats <- interactions %>%
  mutate(p_val_adj = p.adjust(p_val, method = "fdr")) %>%
  left_join(spatial_long, by = c("cell_type1", "cell_type2")) %>%
  filter(!is.na(spatial_corr), spatial_corr > 0.15) %>%
  filter(p_val_adj < 0.05, value > 0) %>%
  group_by(cell_type1) %>%
  ungroup()

top_stats_f <- top_stats %>%
  group_by(cell_type1) %>%
  top_n(2, value) %>%
  ungroup()
pdf(paste0(wd, "pat4_celltalker_circos_cor0.15_2interactions.pdf"), width = 15, height = 13)
circos_plot(ligand_receptor_frame=top_stats_f,
  cell_group_colors=cols_0.15,
  ligand_color="blue",
  receptor_color="red", 
  cex_outer=1.5,
  cex_inner=0.5
)
dev.off()
```

```{r}
cols_0.2 <- c("#8E44AD", "#2171B5", "#C6DBEF", "#B20000", "#FF9900", "#FF6347", 
"#8B4513", "#FFE066", "#c4a310ff", "#27AE60", "#82E0AA", "#800000", "#A93226", 
"#145A32", "#08519C", "#6BAED6")

cols_0.15 <- c("#8E44AD", "#2171B5", "#C6DBEF", "#B20000", "#FF9900", "#FF6347", 
"#8B4513", "#FFE066", "#c4a310ff", "#27AE60", "#82E0AA", "#4682B4", "#1E90FF", 
"#800000", "#A93226", "#990F02", "#B22222", 
"#145A32", "#08519C", "#B0B0B0", "#1CA9C9", "#6BAED6", "#003366")
```

### alternative plot 
```{r}
library(CCPlotR)
cc <- top_stats %>%
  select(source = cell_type1,
         target = cell_type2,
         interaction,
         score = value) %>%
  separate(interaction, into = c("ligand", "receptor"), sep = "_") %>%
  select(source, target, ligand, receptor, score)
````

```{r}
# Idents(pat4_sn) <- "subpopulation"
# exp <- AverageExpression(pat4_sn, group.by = "subpopulation", assays = "RNA", slot = "data")$RNA
# exp <- as.data.frame(exp) %>%
#     tibble::rownames_to_column("gene") %>%
#     pivot_longer(-gene, names_to = "cell_type", values_to = "mean_exp") %>%
#     select(cell_type, gene, mean_exp)
```

```{r}
cell_cols <- c(
    # Immune
    "T-cells-CD8-effector"       = "#4682B4",
    "T-cells-CD4-Treg"           = "#1E90FF",
    "T-cells-atypical"           = "#003366",
    "B-cells-activated"          = "#6bd6d4ff",
    "B/T-mixed-patient4"         = "#C6DBEF",
    "Plasma-cells"               = "#1CA9C9",
    "pDCs"                       = "#08519C",
    "Macrophages-monocyte-derived" = "#003780ff",

    # CAFs
    "mCAF-remodeling"            = "#990F02",
    "mCAF-plastic"               = "#B22222",
    "mCAF-patient4-specific"     = "#A93226",
    "mCAF-classical"             = "#800000",
    "iCAF-classical"             = "#f6d46cff",
    "iCAF-secretory"             = "#c4a310ff",
    "apCAF-classical"            = "#8E44AD",

    # Epithelium
    "Myoepithelial"              = "#27AE60",
    "LumProgenitors-transitional-pat4" = "#82E0AA",
    "LumProgenitors-Cycl-pat4"   = "#145A32",
    "Cancer-Transitional-pat4"   = "#FF6347",
    "Cancer-Basal/Plastic-pat4"  = "#FF9900",
    "Cancer-Basal/Cycl-pat4"     = "#B20000",

    # Endothelial
    "Endothelial"                = "#de88a9ff"
)

cc$source <- factor(cc$source, levels = names(cell_cols))
cc$target <- factor(cc$target, levels = names(cell_cols))
````

#### heatmap interaction cell states
```{r}
pdf(paste0(wd, "pat4_heatmap.pdf"), width = 8, height = 8)
cc_heatmap(cc) + theme(axis.text.x = element_text(angle = 270, vjust = 0.5, hjust = 0))
dev.off()
```

#### dotplot interaction ligand receptor 
```{r}
pdf(paste0(wd, "pat4_dotplot.pdf"), width = 8, height = 12)
cc_dotplot(cc, option = "B", n_top_ints = 100)
dev.off()
```

#### circos
```{r}
pdf(paste0(wd, "pat4_circos.pdf"), width = 15, height = 15)
cc_circos(cc, cell_cols = cell_cols, option = "B", exp_df = exp, n_top_ints = 100)
dev.off()
```

# Check interesting genes in Visium data
```{r}
wd <- paste0(FIGS_ROOT, "/breast_cancer/analysis_cellStateLevel/interesting_genes/")
mergedVis <- readRDS(paste0(rootObj, "merged_visium_deconSubpopulations_simplex.rds"))
```

### C3 pathway interactions
```{r}
filename <- "C3_iCAF_interactions"
genes <- c("C3", "ITGAX", "B2M", "ITGB1", "CXCR4", "LRP1")
cellstates <- c("iCAF-classical", "mCAF-plastic", "pDCs", "B-cells-activated", "B/T-mixed-patient4")

plot_pat4_genes_and_cellstates <- function(pat4, wd, filename, genes, cellstates) {
    # Plot gene features
    DefaultAssay(pat4) <- "Spatial"
    p_genes <- MapFeatures(
        pat4, features = genes, pt_size = 1, ncol = length(genes), drop_na = TRUE, label_by = "sample_id",
        colors = viridis::turbo(11), override_plot_dims = TRUE
    )
    ggsave(
        paste0(wd, filename, "_pat4_genes.jpg"),
        plot = p_genes, height = 5, width = 5 * length(genes), dpi = 400, limitsize = FALSE
    )
    DefaultAssay(pat4) <- "subpopulation"
    p_cellstates <- MapFeatures(
        pat4, features = cellstates, pt_size = 1, ncol = length(cellstates), drop_na = TRUE, label_by = "sample_id",
        colors = viridis::turbo(11), override_plot_dims = TRUE
    )
    ggsave(
        paste0(wd, filename, "_pat4_cellstates.jpg"),
        plot = p_cellstates, height = 5, width = 5 * length(cellstates), dpi = 400, limitsize = FALSE
    )
}
```

#### potential tolerigenic state in immune aggregates induced by C3 
```{r}
filename <- "C3_tolerigenic_state"
cellstates <- c("iCAF-classical", "pDCs", "B-cells-activated", "B/T-mixed-patient4")
genes <- c(
  "IDO1",   # tryptophan metabolism → T cell suppression
  "CD274",  # PD-L1 checkpoint ligand
  "SOCS1",  # dampens cytokine signaling
  "SOCS3",  # negative feedback on pro-inflammatory pathways
  "CTLA4",  # inhibitory receptor on T cells
  "TIGIT",  # inhibitory receptor on T cells
  "CD38"    # regulatory B cell marker
)
plot_pat4_genes_and_cellstates(pat4, wd, filename, genes, cellstates)
```

##### pro-inflammatory state
```{r}
filename <- "proinflammatory_state"
cellstates <- c("iCAF-classical", "pDCs", "B-cells-activated", "B/T-mixed-patient4")
genes <- c(
  "GZMB",    # Granzyme B, cytotoxic T/NK cell effector
  "PRF1",    # Perforin, cytotoxic T/NK cells
  "IFNG",    # interferon gamma, Th1 / cytotoxic cytokine
  "TNF",     # pro-inflammatory cytokine
  "CD8A",    # cytotoxic T cell marker
  "NKG7",    # cytotoxic granule-associated protein
  "CXCL9"    # recruits activated T cells
)
plot_pat4_genes_and_cellstates(pat4, wd, filename, genes, cellstates)
```

### mCAF remodeling 
LAMB1 interaction with mCAF remodeling genes could indicate an initial change in mCAF remodeling state
```{r}
filename <- "mCAF_remodeling_DCIS/mCAF_remodelingDCIS"
genes <- c("LAMB1", "COL5A2", "COL6A1", "VCAN", "THBS2", "ITGAV", "DDR1", "EGFR", "CD47", "CD44", "ITGB1")
cellstates <- c("mCAF-remodeling", "Cancer-Transitional-pat4", "Cancer-Basal/Plastic-pat4")
plot_pat4_genes_and_cellstates(pat4, wd, filename, genes, cellstates)
```

### iCAF secretory and LumProgenitors transitional interactions
LTF-LRP1 interaction could be a driver for the cancer front 
```{r}
filename <- "iCAF_secretory_LumProgTransition/icaf_lumprog"
genes <- c("LTF", "LRP11", "LRP1", "NID1", "DCN", "COL6A3", "PTPRF", "LEPR", "MET", "ITGA1")
cellstates <- c("iCAF-secretory", "LumProgenitors-transitional-pat4")
plot_pat4_genes_and_cellstates(pat4, wd, filename, genes, cellstates)
```

### mCAF populations in the invasive area
Investigating genes responsible for sustaining the invasive growth, invasion and angiogenesis

```{r}
filename <- "mCAF_invasive/mCAF_invasive"
genes <- c("COL5A2", "DDR1", "THBS2", "CD47", "VCAN", "CD44", "ITGB1", "A2M", "LRP1")
cellstates <- c("mCAF-classical", "mCAF-plastic", "mCAF-patient4-specific", "Cancer-Basal/Cycl-pat4")
plot_pat4_genes_and_cellstates(pat4, wd, filename, genes, cellstates)
```

### apCAF interaction with cell states in invasive area
```{r}
filename <- "apCAF/apCAF"
genes <- c("COL5A2", "DDR1", "COL6A2", "ITGB1")
cellstates <- c("apCAF-classical", "mCAF-patient4-specific", "mCAF-classical", "Cancer-Basal/Cycl-pat4")
plot_pat4_genes_and_cellstates(pat4, wd, filename, genes, cellstates)
```

### after checking Xenium plot genes related to immune evasion 
```{r}
filename <- "T_complement"
icaf_module        <- c("CXCL12","CXCL14","CXCL1","CXCL2","IL6","LIF","C3","TGFB1","PDPN","FAP","PDGFRA","COL1A1")
chemotaxis_module  <- c("CXCR4","ACKR3","CXCR3","CXCR6","CCR5","CCR7","S1PR1","ITGAE","ITGB7","CD44")
activation_module  <- c("CD69","FOS","JUN","EGR1","IRF4","CD28","ICOS","TNFRSF9","TNFRSF4")
effector_module    <- c("GZMB","GZMA","PRF1","GNLY","NKG7","IFNG","TNF","IL2","TBX21")
naive_module       <- c("CCR7","SELL","LEF1","TCF7","IL7R")
exhaustion_module  <- c("PDCD1","TOX","NR4A1","CTLA4","LAG3","TIGIT","HAVCR2","ENTPD1")
treg_module        <- c("FOXP3","IL2RA","IKZF2","TNFRSF18","TGFB1","SMAD3","STAT3")
complement_module  <- c("C3AR1","C5AR1","BATF","SOCS1","NR4A1")
genes <- complement_module

DefaultAssay(mergedVis) <- "Spatial"
p <- MapFeatures(
    mergedVis, features = genes, pt_size = 0.6, ncol = length(genes), drop_na = TRUE, label_by = "sample_id",
    colors = viridis::turbo(11), override_plot_dims = TRUE
)
ggsave(
    paste0(wd, "iCAF_trapping_hypothesis/", filename, ".jpg"),
    plot = p, height = 4*8, width = 4 * length(genes), dpi = 400, limitsize = FALSE
)
```

# Looking for FDCs
```{r}
wd <- paste0(FIGS_ROOT, "/breast_cancer/analysis_cellStateLevel/fdc_frc_analysis/")
int.all <- readRDS(paste0(DATA_ROOT, "/r_objects/breast_cancer/single_nuclei/cross_patient/integrated_all_subpopulations.rds"))
mergedVis <- readRDS(paste0(DATA_ROOT, "/r_objects/breast_cancer/visium/cross_patient/merged_visium_deconSubpopulations_simplex.rds"))
```

### FDC
```{r}
wd <- paste0(FIGS_ROOT, "/breast_cancer/analysis_cellStateLevel/fdc/")
pos_genes <- c("CXCL13", "FDCSP", "CR1", "CR2", "SRGN", "PAPPA", "MFGE8", "CLU", "SLC1A2")
neg_genes <- c("PTPRC", "KRT18", "PECAM1", "ADIPOQ", "CD79A", "CD3E", "LYZ", "EPCAM", "MS4A1")
genes <- c(pos_genes, neg_genes)
DefaultAssay(int.all) <- "RNA"
# lin <- subset(int.all, subset = lineage == "immune")
lin <- int.all
p <- DotPlot(lin, features = genes, group.by = "subpopulation") + 
        coord_flip() +
        scale_color_gradientn(colors = c("blue", "white", "darkred")) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))
ggsave(paste0(wd, "all_dotplot.pdf"), p, dpi = 400, height = 5, , width = 12.5, limitsize = FALSE)
```

#### select clusters potentially containing FDCs
```{r}
clusters_to_investigate <- c(
  "T-cells_cycling",
  "Myeloid_atypical_patient6",
  "B/T-mixed_patient4",
  "Cancer_Basal_AntigPres_pat5",
  "Cancer_Basal_EMT_pat5",
  "Cancer_Basal_stem-like_pat6",
  "Cancer_Basal_stress_pat6",
  "Cancer_Lum_pat7",
  "Cancer_LumA_pat8"
)
sub <- subset(int.all, subset = subpopulation %in% clusters_to_investigate)
set.seed(1)
sub <- sub %>%
  NormalizeData() %>%
  FindVariableFeatures() %>%
  ScaleData() %>%
  RunPCA() %>%
  RunUMAP(dims = 1:30)
sub <- FindNeighbors(sub, dims = 1:30)
for (res in c(0.1, 0.25, .3, .4, .5, 0.8, 1, 1.25)){
  sub <- FindClusters(sub, resolution = res, algorithm = 1, verbose = FALSE)
}
d1 <- DimPlot(sub, reduction = "umap", group.by = "RNA_snn_res.0.1", label = TRUE) + 
    DimPlot(sub, reduction = "umap", group.by = "RNA_snn_res.0.25", label = TRUE) + 
    DimPlot(sub, reduction = "umap", group.by = "RNA_snn_res.0.3", label = TRUE) + 
    DimPlot(sub, reduction = "umap", group.by = "RNA_snn_res.0.4", label = TRUE)
d2 <- DimPlot(sub, reduction = "umap", group.by = "RNA_snn_res.0.5", label = TRUE) +
    DimPlot(sub, reduction = "umap", group.by = "RNA_snn_res.0.8", label = TRUE) + 
    DimPlot(sub, reduction = "umap", group.by = "RNA_snn_res.1", label = TRUE) + 
    DimPlot(sub, reduction = "umap", group.by = "RNA_snn_res.1.25", label = TRUE)
pdf(paste0(wd, "FDC_candidates_umapRes.pdf"), width = 20, height = 20)
d1 / d2
dev.off()
```

```{r}
sub$seurat_clusters <- sub$RNA_snn_res.0.1
DefaultAssay(sub) <- "RNA"
p <- DotPlot(sub, features = genes, group.by = "seurat_clusters") + 
        coord_flip() +
        scale_color_gradientn(colors = c("blue", "white", "darkred")) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))
ggsave(paste0(wd, "FDC_candidates_dotplot.pdf"), p, dpi = 400, height = 5, limitsize = FALSE)
p <- FeaturePlot(sub, features = genes)
ggsave(paste0(wd, "FDC_candidates_featureplot.pdf"), p, dpi = 400, height = 20, width = 20, limitsize = FALSE)
p <- DimPlot(sub, group.by = c("seuratAnnotation_major", "subpopulation"))
ggsave(paste0(wd, "FDC_candidates_identity.pdf"), p, dpi = 400, height = 5, width = 13, limitsize = FALSE)
```

#### separate signal from other cells and FDCs
```{r}
fdc <- subset(sub, subset = (FDCSP > 1 | CXCL13 > 1) & MS4A1 < 1 & LYZ < 1 & ITGAX < 1 & EPCAM < 1)
set.seed(1)
fdc <- fdc %>%
  NormalizeData() %>%
  FindVariableFeatures() %>%
  ScaleData() %>%
  RunPCA() %>%
  RunUMAP(dims = 1:30)
fdc <- FindNeighbors(fdc, dims = 1:30)
for (res in c(0.1, 0.25, .3, .4, .5, 0.8, 1, 1.25)){
  fdc <- FindClusters(fdc, resolution = res, algorithm = 1, verbose = FALSE)
}
d1 <- DimPlot(fdc, reduction = "umap", group.by = "RNA_snn_res.0.1", label = TRUE) + 
    DimPlot(fdc, reduction = "umap", group.by = "RNA_snn_res.0.25", label = TRUE) + 
    DimPlot(fdc, reduction = "umap", group.by = "RNA_snn_res.0.3", label = TRUE) + 
    DimPlot(fdc, reduction = "umap", group.by = "RNA_snn_res.0.4", label = TRUE)
d2 <- DimPlot(fdc, reduction = "umap", group.by = "RNA_snn_res.0.5", label = TRUE) +
    DimPlot(fdc, reduction = "umap", group.by = "RNA_snn_res.0.8", label = TRUE) + 
    DimPlot(fdc, reduction = "umap", group.by = "RNA_snn_res.1", label = TRUE) + 
    DimPlot(fdc, reduction = "umap", group.by = "RNA_snn_res.1.25", label = TRUE)
pdf(paste0(wd, "FDC_cleaned_umap.pdf"), width = 10, height = 10)
d1 / d2
dev.off()
fdc$seurat_clusters <- fdc$RNA_snn_res.0.1
p <- DotPlot(fdc, features = genes, group.by = "seurat_clusters") + 
        coord_flip() +
        scale_color_gradientn(colors = c("blue", "white", "darkred")) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))
ggsave(paste0(wd, "FDC_clean_dotplot.pdf"), p, dpi = 400, height = 5, limitsize = FALSE)
p <- FeaturePlot(fdc, features = genes)
ggsave(paste0(wd, "FDC_clean_featureplot.pdf"), p, dpi = 400, height = 20, width = 20, limitsize = FALSE)
# marker genes 
Idents(fdc) <- "seurat_clusters"
markers <- FindAllMarkers(fdc, assay = "RNA")
top_markers <- markers %>%
    group_by(cluster) %>%
    top_n(n = 5, wt = avg_log2FC)
p <- DotPlot(fdc, features = unique(top_markers$gene)) + # the subset is to fix a bug related to NAs
        coord_flip() +
        scale_color_gradientn(colors = c("blue", "white", "darkred")) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))
ggsave(
    filename = paste0(wd, "FDC_clean_markers_dotplot.pdf"),
    plot = p,   
    width = 7,
    height = 5
)
```

#### further purify: cluster 0 and 1
```{r}
fdc_final <- subset(fdc, subset = seurat_clusters %in% c(0,1))
set.seed(1)
fdc_final <- fdc_final %>%
  NormalizeData() %>%
  FindVariableFeatures() %>%
  ScaleData() %>%
  RunPCA() %>%
  RunUMAP(dims = 1:30)
fdc_final <- FindNeighbors(fdc_final, dims = 1:30)
for (res in c(0.1, 0.25, .3, .4, .5, 0.8, 1, 1.25)){
  fdc_final <- FindClusters(fdc_final, resolution = res, algorithm = 1, verbose = FALSE)
}
d1 <- DimPlot(fdc_final, reduction = "umap", group.by = "RNA_snn_res.0.1", label = TRUE) + 
    DimPlot(fdc_final, reduction = "umap", group.by = "RNA_snn_res.0.25", label = TRUE) + 
    DimPlot(fdc_final, reduction = "umap", group.by = "RNA_snn_res.0.3", label = TRUE) + 
    DimPlot(fdc_final, reduction = "umap", group.by = "RNA_snn_res.0.4", label = TRUE)
d2 <- DimPlot(fdc_final, reduction = "umap", group.by = "RNA_snn_res.0.5", label = TRUE) +
    DimPlot(fdc_final, reduction = "umap", group.by = "RNA_snn_res.0.8", label = TRUE) + 
    DimPlot(fdc_final, reduction = "umap", group.by = "RNA_snn_res.1", label = TRUE) + 
    DimPlot(fdc_final, reduction = "umap", group.by = "RNA_snn_res.1.25", label = TRUE)
pdf(paste0(wd, "FDC_pure_umap.pdf"), width = 7, height = 10)
d1 / d2
dev.off()
fdc_final$seurat_clusters <- fdc_final$RNA_snn_res.0.25
p <- DotPlot(fdc_final, features = genes, group.by = "seurat_clusters") + 
        coord_flip() +
        scale_color_gradientn(colors = c("blue", "white", "darkred")) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))
ggsave(paste0(wd, "FDC_pure_dotplot.pdf"), p, dpi = 400, height = 5, limitsize = FALSE)
p <- FeaturePlot(fdc_final, features = genes)
ggsave(paste0(wd, "FDC_pure_featureplot.pdf"), p, dpi = 400, height = 20, width = 20, limitsize = FALSE)
# marker genes 
Idents(fdc_final) <- "seurat_clusters"
markers <- FindAllMarkers(fdc_final, assay = "RNA")
top_markers <- markers %>%
    group_by(cluster) %>%
    top_n(n = 5, wt = avg_log2FC)
p <- DotPlot(fdc_final, features = unique(top_markers$gene)) + # the subset is to fix a bug related to NAs
        coord_flip() +
        scale_color_gradientn(colors = c("blue", "white", "darkred")) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))
ggsave(
    filename = paste0(wd, "FDC_pure_markers_dotplot.pdf"),
    plot = p,   
    width = 7,
    height = 5
)
```


################################################################# UTILS ################################################################
# h5ad conversions

```{r}
library(sceasy)
library(reticulate)
loompy <- reticulate::import('loompy')
```

#### simplex SN obj
```{r}
# seurat to anndata
sceasy::convertFormat(int.all, from="seurat", to="anndata", assay = "RNA",
                       outFile=paste0(rootObj, 'integrated_all_subpopulations.h5ad'))
```

#### spatial obj
```{r}
# seurat to anndata
sceasy::convertFormat(mergedVis, from="seurat", to="anndata", assay = "subpopulation",
                       outFile=paste0(rootObj, 'merged_visium_deconSubpopulations.h5ad'))

```


### fix image pathways
```{r}
mergedVis@tools$Staffli@imgs <- c(
  "patient1_55um" = paste0(DATA_ROOT, "/spaceranger/breast_cancer/55um/patient1_55um/spatial/tissue_lowres_image.png"),
  "patient2_55um" = paste0(DATA_ROOT, "/spaceranger/breast_cancer/55um/patient2_55um/spatial/tissue_lowres_image.png"),
  "patient4_55um" = paste0(DATA_ROOT, "/spaceranger/breast_cancer/55um/patient4_55um/spatial/tissue_hires_image.png"),
  "patient5_55um" = paste0(DATA_ROOT, "/spaceranger/breast_cancer/55um/patient5_55um/spatial/tissue_hires_image.png"),
  "patient6_55um" = paste0(DATA_ROOT, "/spaceranger/breast_cancer/55um/patient6_55um/spatial/tissue_hires_image.png"),
  "patient7_55um" = paste0(DATA_ROOT, "/spaceranger/breast_cancer/55um/patient7_55um/spatial/tissue_hires_image.png"),
  "patient8_55um" = paste0(DATA_ROOT, "/spaceranger/breast_cancer/55um/patient8_55um/spatial/tissue_hires_image.png"),
  "patient9_55um" = paste0(DATA_ROOT, "/spaceranger/breast_cancer/55um/patient9_55um/spatial/tissue_hires_image.png")
)
mergedVis <- LoadImages(mergedVis)
```

### convert staffli to seuratObj@images
```{r}
mergedVis <- UpdateSeuratFromSemla(mergedVis)
```



