---
title: "SIMPlex — per-sample QC, CellBender loading, and doublet removal"
subtitle: "Manuscript figures: Extended Data Fig. 2 (BC QC), Extended Data Fig. 3 (prostate QC)"
author: "Marcos Tito Machado"
output: html_document
---

## Set up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r config, include=FALSE}
# DATA_ROOT, HDF5 library, palettes — single source of truth for paths.
source(here::here("config.R"))
```

```{r}
set.seed(1)
```

## Load libraries
PLEASE LOAD SEURAT4 ENVIRONMENT FOR DOUBLETFINDER TO WORK
```{r package, message='hide',warning='hide',results='hold'}
suppressMessages(require(Seurat))
suppressMessages(require(Matrix))
suppressMessages(require(tidyverse))
suppressMessages(require(patchwork))
suppressMessages(require(ggplot2))
suppressMessages(require(DoubletFinder))
suppressMessages(require(scCustomize))
suppressMessages(require(harmony))

options(future.globals.maxSize = 3.5 * 1024^3)  # 3.5 GiB
```

# Iterate through the samples, filter the data based on cellbender and return seurat objects 
```{r}
samples <- c("3962_1", "3962_2")

for (sample in samples){
  patient <- sub("_.*", "", sample)

  # pathing
  path_cellbender <- paste0(paste0(CELLBENDER, "/breast_cancer/", patient, "/", sample, "_cellbender_filtered.h5")
  path_cellranger <- paste0(paste0(CELLBENDER, "/breast_cancer/", patient, "/", sample, "_filtered_feature_bc_matrix.h5")

  # load cellbender data and create seurat object
  cellbender <- Read_CellBender_h5_Mat(file_name = path_cellbender, use.names = TRUE)
  cbd <- suppressWarnings(CreateSeuratObject(cellbender,  project = "SIMPlex_FFPE_VisHD"))
  cbd$sample <- sample
  rm(cellbender)

  object <- cbd
  # # filter based on 10x CellRanger fixed filtering
  # cellranger <- Seurat::Read10X_h5(filename = path_cellranger, use.names = TRUE)
  # cellranger <- CreateSeuratObject(cellranger,  project = "SIMPlex_FFPE_Xen")
  # cellranger$sample <- sample
  # features <- rownames(cellranger)
  # object <- subset(cbd, features = features)
  # rm(cellranger, features)

  # print(paste0(as.character(length(rownames(cbd)) - length(rownames(object))), paste0(" features removed; sample ", sample)))
  saveRDS(object, paste0(paste0(SN_RDS, "/breast_cancer/per_sample/"), patient, "/", sample, ".rds"))
}
```

## Merge samples 
```{r}
samplePath <- paste0(SN_RDS, "/breast_cancer/per_sample/")
spls <- list()
for (splname in samples) {
    patient <- sub("_.*", "", sample)
    spl <- readRDS(paste0(samplePath, patient, "/", splname, ".rds"))
    spl@meta.data$sample <- splname
    spls[[splname]] <- spl
}
merged <- merge(x = spls[[1]], y = spls[-1])
table(merged@meta.data$sample) #number of cells from each pool
saveRDS(merged, paste0(samplePath, patient, "/", patient, "_merged.rds"))
spl <- merged
```

### sample by sample
(if doing not integrated)
```{r}
splname <- samples[3]
spl <- readRDS(paste0(samplePath, splname, ".rds"))
```

### Wipe old information 
```{r}
spl <- merged
DefaultAssay(spl) <- "RNA"
toKeep <- c("sample", "nCount_RNA", "nFeature_RNA", "percent.mt", "orig.ident")
spl@meta.data <- spl@meta.data[, toKeep, drop = FALSE]
spl[["SCT"]] <- NULL
spl@meta.data$sample_id <- splname
Idents(spl) <- "sample_id"
```

## QC plot 

```{r}
spl$percent.mt <- PercentageFeatureSet(spl, pattern = "^MT-")
feats <- c("nFeature_RNA", "nCount_RNA", "percent.mt")

p1 <- VlnPlot(spl, features = feats, pt.size = 0, ncol = 3) + NoLegend()
p2 <- ggplot() +
  geom_histogram(data = spl[[]], aes(nFeature_RNA), fill = "red", alpha = 0.7, bins = 200) +
  ggtitle("Unique genes per spot") + geom_vline(xintercept = 50, color = "black", linetype = "dashed") + coord_flip()
c1 <- p1 - p2
ggsave(paste0(paste0(FIGS_ROOT, "/breast_cancer/qc_doubletRemoval/", patient, "/", patient,"_qc.pdf"), c1, width = 30, height = 10)
```


## Filtering (number of cells removed) on spl object

```{r qc_filtering}
selected_c <- WhichCells(spl, expression = nFeature_RNA > 50)
selected_f <- rownames(spl)[Matrix::rowSums(spl) > 3]
spl.filt <- subset(spl, features = selected_f, cells = selected_c)

length(colnames(spl)) - length(colnames(spl.filt)) #number of cells removed after filtering
rm(spl)
gc()
```


## Normalize

```{r norm}
set.seed(1)
library(glmGamPoi)
options(future.globals.maxSize = 4 * 1024^3)
spl.filt <- SCTransform(spl.filt, vst.flavor = "v2", verbose = FALSE)
```


## PCA

```{r pca}
set.seed(1)
spl.filt <- RunPCA(spl.filt, features = VariableFeatures(object = spl.filt), verbose = FALSE)
```


## Dimensionality check
```{r elbow_plot_1}
p <- ElbowPlot(spl.filt, reduction = "pca", ndims = 50)
ggsave(paste0(paste0(FIGS_ROOT, "/breast_cancer/qc_doubletRemoval/", patient, "/clustering/", patient, "_elbow.pdf"), p, width = 10, height = 10)
```


## UMAP
```{r umap}
set.seed(1)
spl.filt <- RunUMAP(spl.filt, dims = 1:30, verbose = FALSE)
```


## Plot UMAP

```{r}
Idents(spl.filt) <- "sample"
wd <- paste0(paste0(FIGS_ROOT, "/breast_cancer/qc_doubletRemoval/", patient, "/clustering/")

pdf(paste0(wd, patient, "_pca_before_doublets_removal.pdf"), width = 10, height = 10)
DimPlot(spl.filt, reduction = 'pca', label = TRUE) + ggtitle("spl SN", "pca_before_doublets_removal")
dev.off()

pdf(paste0(wd, patient, "_umap_before_doublets_removal.pdf"), width = 10, height = 10)
DimPlot(spl.filt, reduction = 'umap', label = TRUE) + ggtitle("spl SN", "umap_before_doublets_removal")
dev.off()

# pdf(paste0(wd, splname, "_umap_before_doublets_removal_by_sample.pdf"), width = 10, height = 10)
# DimPlot(spl.filt, reduction = 'umap', group.by = "sample_id", label = TRUE) + ggtitle("spl SN", "umap_before_doublets_removal_by_sample")
# dev.off()
```

# Save n load
```{r}
saveRDS(spl.filt, paste0(paste0(SN_RDS, "/breast_cancer/per_sample/"), patient, "/", patient, "_merged.rds"))
```
```{r}
spl.filt <- readRDS(paste0(paste0(SN_RDS, "/breast_cancer/per_sample/"), patient, "/", patient, "_merged.rds"))
```

## Predict doublets
### Apply on list of objects 

```{r doubletfinder}
wd <- paste0(paste0(FIGS_ROOT, "/breast_cancer/qc_doubletRemoval/", patient, "/doublets/")
if (!dir.exists(wd)) {
  dir.create(wd, recursive = TRUE)
}
data_list <- list()
datasets <- unique(spl.filt$sample_id)
# Filter datasets to include only the specified sample(s)
toDoublet <- splname
datasets <- datasets[datasets %in% splname]

data_list <- lapply(datasets, function(x) {
  x <- as.character(x)
  subset <- spl.filt[, spl.filt@meta.data$sample_id == x]

  # Run parameter optimization with paramSweep
  sweep.res <- paramSweep_v3(subset, PCs = 1:30, sct = TRUE)

  sweep.stats <- summarizeSweep(sweep.res, GT = FALSE)
  bcmvn <- find.pK(sweep.stats)
  pdf(paste0(wd, "pK_estimation_optimized_individual_datasets_prefiltered_adjusted_", x, "_pK_barchart.pdf"), width = 10, height = 10)
  barplot(bcmvn$BCmetric, names.arg = bcmvn$pK, las=2, main = paste0(x, " barchart_before_doublets_removal"))
  dev.off()
  pK <- bcmvn$pK[which.max(bcmvn$BCmetric)]
  BC_pK <- as.numeric(as.character(pK[[1]])) #note: use a single numeric value as pK parameter in the doubletFinder_v3()

  pdf(paste0(wd, "pK_estimation_optimized_individual_datasets_prefiltered_adjusted_", x, "_pK.pdf"), width = 10, height = 10)
  print(ggplot(bcmvn, aes(pK, BCmetric, group = 1)) + geom_point() + geom_line())
  dev.off()

  # define the expected number of doublet cellscells.
  nExp <- round(ncol(subset) * 0.08) # expect 8% doublets
  subset <- doubletFinder_v3(subset, pN = 0.25, pK = BC_pK, nExp = nExp, PCs = 1:30, sct = TRUE)

  subset@meta.data$pANN <- subset@meta.data[, length(colnames(subset@meta.data)) - 1]
  subset@meta.data[, length(colnames(subset@meta.data)) - 2] <- NULL
  subset@meta.data$DF.classifications <- subset@meta.data[, length(colnames(subset@meta.data)) - 1]
  subset@meta.data[, length(colnames(subset@meta.data)) - 2] <- NULL

  return(subset)
})
```


## Save list of objects with doublets found

```{r}
saveRDS(data_list, paste0(paste0(SN_RDS, "/breast_cancer/per_sample/"), patient, "/", patient, "_merged_postDoublet.rds"))
```

```{r}
spl.doub <- data_list[[1]]
```


## Doublet removal
```{r filter.doublets}
colnames(spl.doub@meta.data)
spl.filt_noDoublets <-  spl.doub[, spl.doub@meta.data[, "DF.classifications"] == "Singlet"]
```

```{r}
length(colnames(spl.filt)) - length(colnames(spl.filt_noDoublets)) #number of doublets removed
```

## Normalize, highly variable features, scaling and PCA

```{r pre-proc}
set.seed(1)
spl.filt_noDoublets <- SCTransform(spl.filt_noDoublets, vst.flavor = "v2", verbose = FALSE)
spl.filt_noDoublets <- FindVariableFeatures(spl.filt_noDoublets, selection.method = "vst", verbose = FALSE)
all.genes <- rownames(spl.filt_noDoublets)
spl.filt_noDoublets <- ScaleData(spl.filt_noDoublets, features = all.genes, verbose = FALSE)
spl.filt_noDoublets <- RunPCA(spl.filt_noDoublets, features = VariableFeatures(object = spl.filt_noDoublets), verbose = FALSE)
```

## Dimensionality check
```{r elbow_plot_2}
p <- ElbowPlot(spl.filt_noDoublets, reduction = "pca", ndims = 50)
ggsave(paste0(paste0(FIGS_ROOT, "/breast_cancer/qc_doubletRemoval/", patient, "/clustering/", patient, "elbow_noDoublets.pdf"), p, width = 10, height = 10)
```

## UMAP
```{r umap_sample}
set.seed(1)
Idents(spl.filt) <- "sample"
wd <- paste0(paste0(FIGS_ROOT, "/breast_cancer/qc_doubletRemoval/", patient, "/clustering/")
spl.filt_noDoublets <- RunUMAP(spl.filt_noDoublets, dims = 1:30, verbose = FALSE)

pdf(paste0(wd, patient, "_pca_after_doublets_removal.pdf"), width = 10, height = 10)
DimPlot(spl.filt_noDoublets, reduction = "pca", group.by = "sample", label = TRUE) + ggtitle("pca_after_doublet_removal")
dev.off()

pdf(paste0(wd, patient, "_umap_after_doublets_removal.pdf"), width = 10, height = 10)
DimPlot(spl.filt_noDoublets, reduction = "umap", group.by = "sample", label = TRUE) + ggtitle("umap_after_doublet_removal")
dev.off()
```

## Batch effect correction w/ Harmony 
Only done for replicates / integrated dataset

```{r}
options(repr.plot.height = 2.5, repr.plot.width = 6)
spl.filt_noDoublets <- spl.filt_noDoublets %>% 
    RunHarmony("sample", plot_convergence = TRUE)
```

## UMAP after harmony
Only done for replicates / integrated dataset

```{r umap_sample}
set.seed(1)
Idents(spl.filt) <- "sample"
wd <- paste0(paste0(FIGS_ROOT, "/breast_cancer/qc_doubletRemoval/", patient, "/clustering/")
spl.filt_noDoublets <- RunUMAP(spl.filt_noDoublets, dims = 1:30, verbose = FALSE, reduction = "harmony")

pdf(paste0(wd, "pca_after_harmony.pdf"), width = 10, height = 10)
DimPlot(spl.filt_noDoublets, reduction = "harmony", group.by = "sample", label = TRUE) + ggtitle("pca_after_harmony")
dev.off()

pdf(paste0(wd, "umap_after_harmony.pdf"), width = 10, height = 10)
DimPlot(spl.filt_noDoublets, reduction = "umap", group.by = "sample", label = TRUE) + ggtitle("umap_after_harmony")
dev.off()
```


## Find neighbours & Cluster
```{r neighbours_cluster}
set.seed(1)
Idents(spl.filt) <- "sample"
wd <- paste0(paste0(FIGS_ROOT, "/breast_cancer/qc_doubletRemoval/", patient, "/clustering/")
spl.filt_noDoublets <- FindNeighbors(spl.filt_noDoublets, dims = 1:30, verbose = FALSE)

for (res in c(0.1, 0.25, .5, 0.8, 1, 1.25)){
  spl.filt_noDoublets <- FindClusters(spl.filt_noDoublets, resolution = res, algorithm = 1, verbose = FALSE)
}

d1 <- DimPlot(spl.filt_noDoublets, reduction = "umap", group.by = "SCT_snn_res.0.1", label = TRUE) + 
  DimPlot(spl.filt_noDoublets, reduction = "umap", group.by = "SCT_snn_res.0.25", label = TRUE) + 
  DimPlot(spl.filt_noDoublets, reduction = "umap", group.by = "SCT_snn_res.0.5", label = TRUE)
d2 <- DimPlot(spl.filt_noDoublets, reduction = "umap", group.by = "SCT_snn_res.0.8", label = TRUE) + 
  DimPlot(spl.filt_noDoublets, reduction = "umap", group.by = "SCT_snn_res.1", label = TRUE) + 
  DimPlot(spl.filt_noDoublets, reduction = "umap", group.by = "SCT_snn_res.1.25", label = TRUE)

pdf(paste0(wd, patient, "_umap_cluster_resolutions.pdf"), width = 30, height = 15)
d1 / d2
dev.off()
```

## plot 3D UMAP

```{r}
library(plotly)
library(htmlwidgets)

# Run UMAP in 3D
spl.filt_noDoublets <- RunUMAP(spl.filt_noDoublets, reduction = "harmony", dims = 1:30, n.components = 3)

# Extract UMAP embeddings
umap_3d <- Embeddings(spl.filt_noDoublets, "umap")

# Create a data frame for plotting
umap_3d_df <- as.data.frame(umap_3d)
umap_3d_df$cell <- rownames(umap_3d_df)
umap_3d_df$cluster <- Idents(spl.filt_noDoublets)

# Plot using plotly 
p <- plot_ly(umap_3d_df, x = ~UMAP_1, y = ~UMAP_2, z = ~UMAP_3, color = ~cluster, text = ~cell, colors = "Set1") %>%
  add_markers(marker = list(size = 3)) %>%
  layout(scene = list(xaxis = list(title = 'UMAP 1'),
                      yaxis = list(title = 'UMAP 2'),
                      zaxis = list(title = 'UMAP 3')),
         title = "3D UMAP Plot")

# Save the plot as an HTML file
saveWidget(p, file = paste0(FIGS_ROOT, "/breast_cancer/qc_doubletRemoval/pre_annotation_singleNuclei/clustering/umap_harmony_3D.html"))
```


## Save object
```{r save_RDS}
saveRDS(spl.filt_noDoublets, file = paste0(paste0(SN_RDS, "/breast_cancer/per_sample/"), patient, "/", patient, "_noDoublets.rds"))
```

## Session info

```{r session_info}
sessionInfo()
```
